
# Should be just repeat plots for Figure 4 (simulated gene expression data)

use branch 'alternative_algorithm'
# Import package

```{r}
# path to save everything
dir_to_save_fig <-  "../../out/simulated_gene_expression_alternative/"
dir.create(file.path(".", dir_to_save_fig), showWarnings = F, recursive = T)
unloadNamespace("DualSimplex")
devtools::load_all(path='../../../linseed2/')
```
# Dependencies
```{r}
#rasterize points to reduce size of plots
options("linseed-rasterize"=T)
```



```{r}
library(reshape2)
library(ggplot2)
library(dplyr)
library(ggrepel)
library(ggrastr)
library(RColorBrewer)
```

# Train model on simulated data

## Create object

```{r fig.height = 5, fig.width = 11}
n_ct <- 3

set.seed(3)
sim <- create_simulation(n_genes = 10000,
                         n_samples = 100,
                         n_cell_types = n_ct,
                         with_marker_genes = FALSE)
sim <- sim %>% add_noise(noise_deviation = 0.2)

data_raw <- sim$data
true_basis <- sim$basis
true_proportions <- sim$proportions



dso <- DualSimplexSolver$new()

dso$set_data(data_raw)

plane_distance_threshold = 0.05 # Change here several times to see result, start with big and lower it
dso$project(3)
dso$distance_filter(plane_d_lt = plane_distance_threshold, zero_d_lt = NULL, genes = T)

dso$project(3)
dso$plot_svd_history()
dso$init_solution("random")
dso$plot_projected("zero_distance", "zero_distance",with_solution=TRUE, use_dims = list(2:3))
```
## Make specific init, which was in a paper
```{r}
set.seed(15)
dso$init_solution("random")
dso$plot_projected("plane_distance", "plane_distance",with_solution=TRUE, use_dims = list(2:3))

```


## Make 5 steps of optimization
```{r}
blocks <- 5
iterations <- 2000
coef_hinge_H <-  0.1
coef_hinge_W <-  0.1


for (i in 1:blocks) {
  
  dso$optim_solution(round(iterations/blocks), 
                     optim_config(coef_hinge_H = coef_hinge_H,
                                  coef_hinge_W = coef_hinge_W,
                                  coef_der_X = 0.1, 
                                  coef_der_Omega = 0.1
                                  ))
  curr_X <- dso$st$solution_proj$X  # this is how we can extract solution on a fly
  curr_Omega <- dso$st$solution_proj$Omega # this is how we can extract solution on a fly
  coef_hinge_H <-  coef_hinge_H / 2
  coef_hinge_W <-  coef_hinge_W / 2
}
```

```{r fig.height = 5, fig.width = 11}
dso$plot_projected("zero_distance", "zero_distance",with_solution=TRUE, use_dims = list(2:3))
```

```{r}
dso$plot_error_history()
```




## This is how textract metadata (terms for equations)
```{r}
solution_proj <- dso$st$solution_proj
proj <-  dso$st$proj

solution <- dso$finalize_solution()
N <- dso$st$proj$meta$N
M <- dso$st$proj$meta$M
X<- solution_proj$X  
R <-  proj$meta$R
S <-  dso$st$proj$meta$S
Omega <- t(dso$st$solution_proj$Omega)
ones_S <- as.matrix((rep(1, dim(S)[[2]])))
ones_R <- as.matrix((rep(1, dim(R)[[2]])))
ones_K <- as.matrix((rep(1, 3)))
```

## This is how to save the model
```{r}
dso$set_save_dir(paste0(dir_to_save_fig,"/simulations_fig_4_save"))
dso$save_state()
```

## This is how to load model
```{r}
save_dir <- paste0(dir_to_save_fig,"/simulations_fig_4_save")
dso <- DualSimplexSolver$from_state(save_dir)

# Note: this wasn't saved, it's not included in state,
# so we have to set it again
dso$set_display_dims(list(NULL, 2:3)) 
n_ct <- dso$st$n_cell_types
```

## Extract 5 uniformly distributed points from the training log
```{r}
solution_start_end <- get_solution_history(dso$st$solution_proj, iterations/blocks)
X_hist <- as.data.frame(solution_start_end$X)
Omega_hist <- as.data.frame(solution_start_end$Omega)

colnames(X_hist) <- c("X","Y","Z", "k","iteration")
colnames(Omega_hist) <- c("X","Y","Z", "k","iter")
X_hist$i <-  rep(0:blocks, each=3)
Omega_hist$i <-  rep(0:blocks, each=3)
```


# Save error how it appears in the paper

```{r}
plotErrorsWithRastr = function(metadata, variables = c("deconv_error","lamdba_error","beta_error",
"D_h_error","D_w_error","total_error")) {
  solution_proj <- metadata$st$solution_proj
  error_statistics <- solution_proj$optim_history$errors_statistics
  
  toPlot <- data.frame(error_statistics[, variables] )

  toPlot$iteration <- 0:(nrow(error_statistics)-1)
  toPlot <- melt(toPlot,id.vars="iteration",measure.vars = variables)
  plt <- ggplot(toPlot,aes(x=iteration,y=log10(value),color=variable)) +
    # rasterise(geom_point(size=0.2),dpi=600) +
     rasterise(geom_line(size=0.8), dpi=600) + theme_minimal() + labs(color="Errors")
  return(plt)
    }

errors_plot <- plotErrorsWithRastr(dso, variables = c("deconv_error",
                                 "lamdba_error",
                                 "beta_error",
                                 "total_error")) + 
   theme_minimal(base_size=14,base_family = 'sans')

#ylim(-50, NA)
# measure_points <-   data.frame( iteration = unique(X_hist$iter), value=10)
# 
# 
#   errors_plot <- errors_plot + geom_label_repel(data = measure_points, size = 5,
#                                 mapping = aes(label=i), colour="white")      #label


filename <- paste0(dir_to_save_fig, "simulation_trajectory", ".svg")
ggsave(file=filename, plot=errors_plot, width=5, height=3, device=svglite)
errors_plot
```


## Triangles as how they are in paper


```{r}

plot_gradient <- function(toPlot, endpoints, colors) {
  blocks <- max(endpoints$i)
  print(blocks)
  
  plt <- ggplot(toPlot, aes(x=Y, y=Z)) +
     rasterise(geom_point(color=colors[4],size=1,alpha=0.8), dpi=600)
  for (j in 0:(blocks-1)) {
    
    for (c in 1:n_ct) {
      plt <- plt + geom_line(data = endpoints %>% 
                                  filter(i %in% c(j,j+1)) %>%
                                  filter(k==c),
                                size=1,
                               color = 'gray44' )  
    }
  }

  plt <- plt + geom_polygon(data=endpoints %>% filter(i==0),
                        size=1, fill=NA, color = colors[7], linetype = "dashed")  # triangle init
  plt <-  plt  + geom_polygon(data=endpoints %>% filter(i==blocks), 
                              size=1, fill=NA, color = colors[6])  # triangle final
  plt <-  plt + geom_point(data=endpoints %>% filter(i!=blocks),
                           fill = colors[5],
                           color = colors[4], size=3, shape = 21, stroke=1)   # trajectory points
  plt <-  plt  + geom_point(data=endpoints %>% filter(i==blocks),
                            fill = colors[6],
                            color = colors[5], size=3, shape = 21, stroke=1) # final result points
  plt <- plt + geom_label_repel(data = endpoints, size = 5,
                                mapping = aes(label=i)) +        #label
    theme_minimal(base_size = 25, base_family = 'sans') + theme(axis.line= element_blank(),
                            text = element_text(size = 16)) 
  plt
}
```





## X
```{r}
dims <- 3
points_col <- brewer.pal(9,"Blues")

## plot X  
toPlot <- as.data.frame(dso$st$proj$X)[,1:dims]
colnames(toPlot) <- c("X","Y","Z")

pltX <- plot_gradient(toPlot, X_hist, points_col)

pltX <- pltX + xlab("R2") + ylab("R3")

filename <- paste0(dir_to_save_fig, "simulation_X", ".svg")
ggsave(file=filename, plot=pltX, width=5, height=4, device=svglite)
pltX
```
## Omega 

```{r}
## plot Omega  

toPlot <- as.data.frame(dso$st$proj$Omega)[,1:dims]
colnames(toPlot) <- c("X","Y","Z")

points_col <- brewer.pal(9,"Oranges")
pltOmega <- plot_gradient(toPlot, Omega_hist, points_col)
pltOmega <- pltOmega + xlab("S2") + ylab("S3")

filename <- paste0(dir_to_save_fig, "simulation_Omega", ".svg")
ggsave(file=filename, plot=pltOmega, width=5, height=4, device=svglite)
pltOmega
```

## Prepare basis/proportions plots
```{r}
solution <- dso$finalize_solution()
names(solution)
solution <- dso$get_solution()
```

```{r fig.width=20, fig.height=5}
ptb <- coerce_pred_true_basis(solution$W, true_basis[rownames(solution$W), ])
ptp <- coerce_pred_true_props(solution$H, true_proportions)
plot_ptp_scatter(ptp)
plot_ptb_scatter(ptb)
```
```{r}
plot_ptp_lines(ptp)
```
