---
title: "Figure 3C: Main Algorithm, Simulation"
output: html_document
---

# Import package
```{r}
source('../R/setup.R')
library(reshape2)
library(ggplot2)
library(dplyr)
library(ggrepel)
library(ggrastr)
library(RColorBrewer)

options("dualsimplex-rasterize" = T)
```


# Train model on simulated data
## Create object
```{r fig.height = 5, fig.width = 11}
n_ct <- 3

set.seed(3)
sim <- create_simulation(n_genes = 10000,
                         n_samples = 100,
                         n_cell_types = n_ct,
                         with_marker_genes = FALSE)
sim <- sim %>% add_noise(noise_deviation = 0.2)

data_raw <- sim$data
true_basis <- sim$basis
true_proportions <- sim$proportions

dso <- DualSimplexSolver$new()
dso$set_data(data_raw)

plane_distance_threshold = 0.05 # Change here several times to see result, start with big and lower it
dso$project(3)
dso$distance_filter(plane_d_lt = plane_distance_threshold,
                    zero_d_lt = NULL,
                    genes = T)

dso$project(3)
dso$plot_svd_history()
dso$init_solution("random")
dso$plot_projected(
  "zero_distance",
  "zero_distance",
  with_solution = TRUE,
  use_dims = list(2:3)
)
```


## Make specific init, which was in a paper
```{r}
set.seed(23)
dso$init_solution("random")
dso$plot_projected(
  "plane_distance",
  "plane_distance",
  with_solution = TRUE,
  use_dims = list(2:3)
)
```


## Make 5000 steps of optimization
```{r}
blocks <- 5
iterations <- 5000


for (i in 1:blocks) {
  dso$optim_solution(
    round(iterations / blocks),
    optim_config(
      coef_hinge_H = 1,
      coef_hinge_W = 1,
      coef_der_X = 0.001,
      coef_der_Omega = 0.001
    )
  )
  curr_X <- dso$st$solution_proj$X  # this is how we can extract solution on a fly
  curr_Omega <- dso$st$solution_proj$Omega # this is how we can extract solution on a fly
}
```

```{r fig.height = 5, fig.width = 11}
dso$plot_projected(
  "zero_distance",
  "zero_distance",
  with_solution = TRUE,
  use_dims = list(2:3)
)
```

```{r}
dso$plot_error_history()
```


## This is how extract metadata (terms for equations)
```{r}
solution_proj <- dso$st$solution_proj
proj <-  dso$st$proj

solution <- dso$finalize_solution()
N <- dso$st$proj$meta$N
M <- dso$st$proj$meta$M
X <- solution_proj$X
R <-  proj$meta$R
S <-  dso$st$proj$meta$S
Omega <- t(dso$st$solution_proj$Omega)
ones_S <- as.matrix((rep(1, dim(S)[[2]])))
ones_R <- as.matrix((rep(1, dim(R)[[2]])))
ones_K <- as.matrix((rep(1, 3)))
```


## This is how to save the model
```{r}
dso$set_save_dir("../out/dualsimplex_save_fig3")
dso$save_state()
```


## This is how to load model
```{r}
dso <- DualSimplexSolver$from_state("../out/dualsimplex_save_fig3")

# Note: this wasn't saved, it's not included in state,
# so we have to set it again
dso$set_display_dims(list(NULL, 2:3)) 
n_ct <- dso$st$n_cell_types
```


## Extract 5 uniformly distributed points from the training log
```{r}
solution_start_end <- get_solution_history(
  dso$st$solution_proj,
  iterations / blocks
)
X_hist <- as.data.frame(solution_start_end$X)
Omega_hist <- as.data.frame(solution_start_end$Omega)

colnames(X_hist) <- c("X", "Y", "Z", "k", "iteration")
colnames(Omega_hist) <- c("X", "Y", "Z", "k", "iter")
X_hist$i <-  rep(0:blocks, each = 3)
Omega_hist$i <-  rep(0:blocks, each = 3)
```


# Save error how it appears in the paper
```{r}
plotErrorsWithRastr = function(metadata,
                               variables = c(
                                 "deconv_error",
                                 "lamdba_error",
                                 "beta_error",
                                 "D_h_error",
                                 "D_w_error",
                                 "total_error"
                               )) {
  solution_proj <- metadata$st$solution_proj
  error_statistics <- solution_proj$optim_history$errors_statistics
  
  toPlot <- data.frame(error_statistics[, variables])
  
  toPlot$iteration <- 0:(nrow(error_statistics) - 1)
  toPlot <-
    melt(toPlot, id.vars = "iteration", measure.vars = variables)
  plt <-
    ggplot(toPlot, aes(
      x = iteration,
      y = log10(value),
      color = variable
    )) +
    # rasterise(geom_point(size=0.2),dpi=600) +
    rasterise(geom_line(size = 0.8), dpi = 600) + theme_minimal() + labs(color =
                                                                           "Errors")
  return(plt)
}

errors_plot <-
  plotErrorsWithRastr(dso,
                      variables = c("deconv_error",
                                    "lamdba_error",
                                    "beta_error",
                                    "total_error")) +
  theme_minimal(base_size = 14, base_family = 'sans')


ggsave(
  file = "../out/3c_errors_trajectory.svg",
  plot = errors_plot,
  width = 5,
  height = 3,
  device = svglite::svglite
)
errors_plot
```


## Triangles as how they are in paper
```{r}
plot_gradient <- function(toPlot, endpoints, colors) {
  blocks <- max(endpoints$i)
  print(blocks)
  
  plt <- ggplot(toPlot, aes(x = Y, y = Z)) +
    rasterise(geom_point(
      color = colors[4],
      size = 1,
      alpha = 0.8
    ), dpi = 600)
  for (j in 0:(blocks - 1)) {
    for (c in 1:n_ct) {
      plt <- plt + geom_line(
        data = endpoints %>%
          filter(i %in% c(j, j + 1)) %>%
          filter(k == c),
        size = 1,
        color = 'gray44'
      )
    }
  }
  
  plt <- plt + geom_polygon(
    data = endpoints %>% filter(i == 0),
    size = 1,
    fill = NA,
    color = colors[7],
    linetype = "dashed"
  )  # triangle init
  plt <-  plt  + geom_polygon(
    data = endpoints %>% filter(i == blocks),
    size = 1,
    fill = NA,
    color = colors[6]
  )  # triangle final
  plt <-  plt + geom_point(
    data = endpoints %>% filter(i != blocks),
    fill = colors[5],
    color = colors[4],
    size = 3,
    shape = 21,
    stroke = 1
  )   # trajectory points
  plt <-  plt  + geom_point(
    data = endpoints %>% filter(i == blocks),
    fill = colors[6],
    color = colors[5],
    size = 3,
    shape = 21,
    stroke = 1
  ) # final result points
  plt <- plt + geom_label_repel(data = endpoints,
                                size = 5,
                                mapping = aes(label = i)) + # label
    theme_minimal(base_size = 25, base_family = 'sans') + theme(axis.line = element_blank(),
                                                                text = element_text(size = 16))
  plt
}
```


## X
```{r}
dims <- 3
points_col <- brewer.pal(9, "Blues")

## plot X
toPlot <- as.data.frame(dso$st$proj$X)[, 1:dims]
colnames(toPlot) <- c("X", "Y", "Z")

pltX <- plot_gradient(toPlot, X_hist, points_col)

pltX <- pltX + xlab("R2") + ylab("R3")

ggsave(
  file = "../out/3c_x_optimization.svg",
  plot = pltX,
  width = 5,
  height = 4,
  device = svglite::svglite
)
pltX
```


## Omega 
```{r}
toPlot <- as.data.frame(dso$st$proj$Omega)[, 1:dims]
colnames(toPlot) <- c("X", "Y", "Z")

points_col <- brewer.pal(9, "Oranges")
pltOmega <- plot_gradient(toPlot, Omega_hist, points_col)
pltOmega <- pltOmega + xlab("S2") + ylab("S3")

ggsave(
  file = "../out/3c_omega_optimization.svg",
  plot = pltOmega,
  width = 5,
  height = 4,
  device = svglite::svglite
)
pltOmega
```


## Prepare basis/proportions plots
```{r}
solution <- dso$finalize_solution()
names(solution)
solution <- dso$get_solution()
```

```{r fig.width=20, fig.height=5}
ptb <- coerce_pred_true_basis(solution$W, true_basis[rownames(solution$W), ])
ptp <- coerce_pred_true_props(solution$H, true_proportions)
plot_ptp_scatter(ptp)
plot_ptb_scatter(ptb)
```

```{r}
plot_ptp_lines(ptp)
```

