---
title: "Figure 5: NMF, Part 3."
output: html_notebook
---


```{r}
dir_to_save_fig <- "../out/4_NMF/"
dir.create(file.path(".", dir_to_save_fig), showWarnings = F, recursive = T)
source('../R/setup.R')
source('../R/nmf_methods.R')
```


# Pictures NMF problem

## Single Run
### Read pictures (main components)
```{r}
set.seed(4)
size <- 128
number_of_pictures_to_select <- 4

pictures_W <- rhdf5::h5read("../data/nmf_images/merged_imgs.h5", "matrix")
pictures_W <- t(pictures_W)[,1:number_of_pictures_to_select]
current_M <- nrow(pictures_W)
current_K <- ncol(pictures_W)
```


### Make N mixed pictures with or without noise)
```{r}
current_N <- 300
globally_generated_pic_data <- make_picture_data_random_mixtures(current_M, current_N, current_K, pictures_W, noise_deviation = 0)
```


### Solve this unmixing problem using DualSimplex algorithm

#### Create object
```{r}
dso <- DualSimplexSolver$new()
dso$set_data(globally_generated_pic_data$noizy_V)
dso$project(current_K)
dso$run_umap()
dso$plot_projected()
```

#### Train model
We perform `LR_DECAY_STEPS` epochs of training with decreasing learning rate (decrease by `lr_decay`)
  Each epoch we run `PARAMETERS_INCREASE_STEPS` steps, where we increase `lambda` and `beta` by `params_increase`
    For each step we do less iterations starting   `RUNS_EACH_STEP` and decreasing this amount by 0.7 each step

```{r}
dso$init_solution("random")

dso$default_optimization()
solution <- dso$finalize_solution()
```


```{r}
dso$plot_projected("plane_distance", "plane_distance", use_dims=list(2:3, 3:4))
```


```{r}
dso$plot_error_history()
ggarrange(dso$plot_negative_basis_change(), dso$plot_negative_proportions_change())
```


#### Try to visualize solution
(This is not mandatory code, just for your visual understanding)
Here we will plot predicted solution points together with real coordinates in a 3D space of Sinkhorn transformed matrices.

Take solution sinkhorn transformed matrices identified (H_inf and W_inf)
```{r}
sol_no_corr <- dso$st$solution_no_corr
W_solution <- sol_no_corr$Dv_inv_W_row
W_solution[W_solution< 0] = 0
H_solution <- sol_no_corr$H_row
H_solution[H_solution < 0] <- 0
```


Take true solution matrices (Without sinkhorn transformation).
```{r}
true_V <- globally_generated_pic_data$V[rownames(W_solution),colnames(H_solution)]
true_V_noizy <- globally_generated_pic_data$noizy_V[rownames(W_solution),colnames(H_solution)]
true_W <- globally_generated_pic_data$true_W[rownames(W_solution),]
true_H <- globally_generated_pic_data$true_H[,colnames(H_solution)]
```


##### Identify coordinates in projected space for true matrices
Here we will need to perform Sinkhorn transformation for matrix V to restore D_w and D_h matrices. It can take some processing time.
We will do this for both noisy and original matrices
```{r}

true_solution_coordinates <-  dso$get_coordinates_from_external_matrices(true_W, true_H)
X_true <- true_solution_coordinates$X
Omega_true <-  t(true_solution_coordinates$Omega)

```


##### Prepare sets of points to visualize
```{r}
library(plotly)
dims <- dso$st$n_cell_types
## Prepare set of points 


### True corners
X <- X_true
Omega <- Omega_true
X <- X[,1:dims]
Omega <- Omega[1:dims,]
colnames(X) <- paste0("R", c(1:dims))
rownames(Omega) <- paste0("S", c(1:dims))


### Solution corners
X_solution <- dso$st$solution_proj$X
Omega_solution <- t(dso$st$solution_proj$Omega)
X_solution <- X_solution[,1:dims]
Omega_solution <- Omega_solution[1:dims,]
colnames(X_solution) <- paste0("R", c(1:dims))
rownames(Omega_solution) <- paste0("S", c(1:dims))

## Just all the points
### X space
toPlotX <- as.data.frame(dso$st$proj$X)[,1:dims]
colnames(toPlotX) <- paste0("R", c(1:dims))
### Omega space
toPlotOmega <- as.data.frame(dso$st$proj$Omega)[,1:dims]
colnames(toPlotOmega) <- paste0("S", c(1:dims))
```


##### Prepare geometrical lines for plotly
```{r}
get_line_between_points <- function(X_p, Omega_p) {
  n <- 2
  l <- rep(list(1:nrow(X_p)), n)
  permutations <- expand.grid(l)
  indices <- c(t(as.matrix(permutations[permutations[,"Var1"] < permutations[, 'Var2'],] )))
  X_points <- X_p[indices,]
  Omega_points <- Omega_p[,indices]
  return(list(X_points=X_points, Omega_points=Omega_points))
}


## Prepare lines to draw between corners true
line_result <- get_line_between_points(X, Omega)
X_points <- line_result$X_points
Omega_points <- line_result$Omega_points

## Prepare lines to draw between corners true
line_result <- get_line_between_points(X_solution, Omega_solution)
X_points_solution <- line_result$X_points
Omega_points_solution <- line_result$Omega_points

# ## Prepare lines to draw between corners corrected
# line_result <- get_line_between_points(X_corrected, Omega_corrected)
# X_points_corrected <- line_result$X_points
# Omega_points_corrected <- line_result$Omega_points
```


```{r}
t <- list(family = "Helvetica",
          size = 30)

fig_X <-
  plot_ly(
    toPlotX,
    width = 500,
    height = 500,
    type = 'scatter3d',
    mode = 'markers'
  ) %>%
  config(toImageButtonOptions = list(
    format = "svg",
    width = 600,
    height = 600
  )) %>%
  # add all dataset points
  add_trace(
    x = ~ R2,
    y = ~ R3,
    z = ~ R4,
    size = 1,
    marker = list(
      symbol = 'circle',
      color = "gray",
      opacity = 0.8,
      line = list(
        opacity = 0.4,
        color = 'black',
        width = 1
      )
    )
  ) %>%
  # add  true corner coordinates
  add_trace(
    x = X_points[, 'R2'],
    y = X_points[, 'R3'],
    z = X_points[, 'R4'],
    mode = 'markers+lines',
    line = list(color = 'rgb(44, 160, 44)', width = 5)
  ) %>%
  # add solution corner coordinates
  add_trace(
    x = X_points_solution[, 'R2'],
    y = X_points_solution[, 'R3'],
    z = X_points_solution[, 'R4'],
    mode = 'markers+lines',
    line = list(color = 'rgb(0, 44, 160)', width = 5)
  )


fig_X <- fig_X %>% layout(showlegend = FALSE,
                          scene = list(
                            xaxis = list(title = 'R2',  titlefont = t),
                            yaxis = list(title = 'R3', titlefont = t),
                            zaxis = list(title = 'R4', titlefont = t)
                          )) %>% config(toImageButtonOptions = list(
                            format = "svg",
                            width = 600,
                            height = 600
                          ))

fig_X
```


```{r}
n <- 2
# l <- rep(list(1:nrow(X)), n)
# permutations <- expand.grid(l)
# indices <- c(t(as.matrix(permutations[permutations[,"Var1"] < permutations[, 'Var2'],] )))
# Omega_points <- Omega[,indices]


fig_Omega <-
  plot_ly(
    toPlotOmega,
    width = 500,
    height = 500,
    type = 'scatter3d',
    mode = 'markers'
  ) %>%
  config(toImageButtonOptions = list(
    format = "svg",
    width = 600,
    height = 600
  )) %>%
  add_trace(
    x = ~ S2,
    y = ~ S3,
    z = ~ S4,
    size = 3,
    marker = list(
      symbol = 'circle',
      size = 5,
      color = "gray",
      line = list(color = 'black', width = 3)
    )
  ) %>%
  add_trace(
    x = Omega_points['S2', ],
    y = Omega_points['S3', ],
    z = Omega_points['S4', ],
    line = list(color = 'rgb(44, 160, 44)', width = 5),
    mode = 'markers+lines'
  )  %>%
  add_trace(
    x = Omega_points_solution['S2', ],
    y = Omega_points_solution['S3', ],
    z = Omega_points_solution['S4', ],
    line = list(color = 'rgb(0, 44, 160)', width = 5),
    mode = 'markers+lines'
  )


fig_Omega <- fig_Omega %>% layout(showlegend = FALSE,
                                  scene = list(
                                    xaxis = list(title = 'S2',  titlefont = t),
                                    yaxis = list(title = 'S3', titlefont = t),
                                    zaxis = list(title = 'S4', titlefont = t)
                                  )) %>% config(toImageButtonOptions = list(
                                    format = "svg",
                                    width = 600,
                                    height = 600
                                  ))

fig_Omega
```


## Multiple Runs

### Define experimentsmethod
```{r}
run_experiment_picture_random_noizy <-
  function(current_K,
           current_M,
           current_N,
           pictures_W,
           n_data_generations = 3,
           start_with_hinge_W = 1,
           start_with_hinge_H = 1,
           LR_DECAY_STEPS = 7,
           RUNS_EACH_STEP = 500,
           PARAMETERS_INCREASE_STEPS = 2,
           lr_x =  0.1,
           lr_omega = 0.1,
           noise_deviation = 3.5,
           noise_nature = "proportional",
           n_runs_for_each_model = 10) {
    data_generated <-
      lapply(c(1:n_data_generations), function(run_index)
        make_picture_data_random_mixtures(current_M, current_N, current_K, pictures_W, noise_deviation))
    
    total_result <-
      lapply(c(1:n_data_generations), function(run_index) {
        print(paste("data gen ", run_index))
        syntetic_data <- data_generated[[run_index]]
        current_res_methods <- list()
        current_res_methods <-
          NMF::nmf(
            syntetic_data$noizy_V,
            current_K,
            list('lee', 'brunet'),
            .options = 'vp10',
            nrun = n_runs_for_each_model
          )
        
        print("Compute NMF method 'HALSacc'")
        halsacc_res <-
          NMF::nmf(
            syntetic_data$noizy_V,
            current_K,
            hNMF::HALSacc,
            nrun = n_runs_for_each_model,
            .options = 'vp10',
            name = "HALSacc"
          )
        #halsacc_res <- nmf(syntetic_data$V, current_K, halsacc_nmf_algorithm, nrun=10, .options='tp1', name="HALSacc")
        current_res_methods[["HALSacc"]] <- halsacc_res
        print("Compute NMF method 'PGNMF'")
        pgnmf_res <-
          NMF::nmf(
            syntetic_data$noizy_V,
            current_K,
            hNMF::PGNMF,
            nrun = n_runs_for_each_model,
            .options = 'vp10',
            name = "PGNMF"
          )
        # # pgnmf_res <- nmf(syntetic_data$V, current_K, pgnmf_nmf_algorithm, nrun=10, .options='tp1', name="PGNMF")
        current_res_methods[["PGNMF"]] <- pgnmf_res
        #print("Compute NMF method 'ALS'")
        als_nmf_res <-
          NMF::nmf(
            syntetic_data$noizy_V,
            current_K,
            als_nmf_algorithm,
            nrun = n_runs_for_each_model,
            .options = 'p1',
            name = "ALS"
          )
        current_res_methods[["ALS"]] <- als_nmf_res
        print("Compute NMF method 'DualSimplex'")
        linseed_res <- NMF::nmf(
          syntetic_data$noizy_V,
          current_K,
          dualsimplex_nmf_algorithm,
          nrun = n_runs_for_each_model,
          .options = 'vp10',
          name = "DualSimplex",
          start_with_hinge_W = start_with_hinge_W,
          start_with_hinge_H = start_with_hinge_H,
          LR_DECAY_STEPS = LR_DECAY_STEPS,
          RUNS_EACH_STEP = RUNS_EACH_STEP,
          PARAMETERS_INCREASE_STEPS = PARAMETERS_INCREASE_STEPS,
          lr_x = lr_x,
          lr_omega = lr_omega
        )
        current_res_methods[["DualSimplex"]] <- linseed_res
        return(current_res_methods)
      })
    
    
    all_results_flat <- unlist(total_result, recursive = F)
    unique_names <- unique(names(all_results_flat))
    per_method_per_run_result <-
      lapply(unique_names, function(cur_nm)
        all_results_flat[names(all_results_flat) == cur_nm])
    names(per_method_per_run_result) <- unique_names
    # distances_to_check <- plot_metrics(res_methods = per_method_per_run_result, data_used=data_generated, title=paste("K =", as.character(current_K), "M =", as.character(current_M),"N =",as.character(current_N)))
    return (list(data_used = data_generated, fit_objects = per_method_per_run_result))
    
  }
```


### Run experiment

```{r}
current_N  <- 100
# For N = 500 use   start_with_hinge_W=0.1 start_with_hinge_H=0.1
# For N = 100 use   start_with_hinge_W=10 start_with_hinge_H=1
noise_nature <- "proportional"

for (noise_deviation in c(0,2)) {
  result_random <-
    run_experiment_picture_random_noizy(
      current_K = current_K,
      current_M = current_M,
      current_N = current_N,
      n_data_generations = 10,
      n_runs_for_each_model =
        10,
      pictures_W = pictures_W ,
      noise_deviation =
        noise_deviation,
      noise_nature = noise_nature,
      start_with_hinge_W =
        10,
      start_with_hinge_H =
        1,
      LR_DECAY_STEPS = 11,
      RUNS_EACH_STEP = 2000,
      PARAMETERS_INCREASE_STEPS = 3,
      lr_x =  0.1
    )
  
  print("Save to file")
  file_name <-
    paste0(
      dir_to_save_fig,
      noise_nature,
      "_random_noise_",
      as.character(noise_deviation),
      '_random_proportions_',
      current_K ,
      'K_',
      current_N,
      'N_',
      current_M,
      'M_random-H_10runs_10init_random.rds'
    )
  saveRDS(result_random, file = file_name)
}

```


```{r}
check_triangles(result_random$fit_objects)
```

```{r, fig.height = 5, fig.width=20}
distances_to_check <- plot_metrics_box(res_methods = result_random$fit_objects, 
                                       data_used=result_random$data_used, 
                                       title = "test_pictures_result", 
                                       folder_to_save_results = dir_to_save_fig)
distances_to_check$plot 
distances_to_check$W_plot
distances_to_check$H_plot
```

