---
title: "Figure 5: NMF, Part 3."
output: html_notebook
---

```{r}
# path to results and to save everything
dir_to_save_fig <- "../out/4_NMF/"
dir.create(file.path(".", dir_to_save_fig), showWarnings = F, recursive = T)
source("../R/setup.R")
source("../R/nmf_methods.R")
source("../R/figure_utils.R")
```


# Analyze saved NMF results
```{r}
current_K <- 4
current_M <- 16384
current_N <- 100

noise_deviation <- 2
noise_nature <- "proportional" # proportional additive
proportions <- "random_proportions" # simplex random_proportions
```


# Randomly generated proportions

## Read dual simplex objects
```{r}
file_name <- paste0(
  dir_to_save_fig,
  noise_nature,
  '_random_noise_',
  noise_deviation,
  '_',
  proportions,
  '_',
  current_K ,
  'K_',
  current_N,
  'N_',
  current_M,
  'M_random-H_10runs_10init_random.rds'
)

result_random <- readRDS(file = file_name)
```

## Check which methods we have here
```{r}
# remove backslash from name
current_names <- names(result_random$fit_objects)
if ("snmf/l" %in% current_names) {
  current_names[current_names == "snmf/l"] <- "snmf_l"
  names(result_random$fit_objects) <- current_names
  result_random$fit_objects$snmf_l <-
    lapply(result_random$fit_objects$snmf_l, function(result) {
      result@method <- 'snmf_l'
      return(result)
    })
  
}

print(current_names)
```



### Draw result pictures obtained by each method
```{r fig.height = 5, fig.width = 5}

individual_pictures <- T # plot single picture for a file or all K pictures

if (individual_pictures) {
  orig_pictures <- plot_single_pictures(result_random$data_used[[2]]$true_W, "original_picture", dir_to_save_fig)
} else {

orig_pictures <- plot_pictures(result_random$data_used[[2]]$true_W,  "original_pictures", dir_to_save_fig)
}

per_method_pictures <- plot_pictures_for_result(result_random$fit_objects, 
                                                model_index = 1, 
                                                dir_to_save_fig, 
                                                individual_pictures = individual_pictures)
for (k in 1:current_K) {
  image(per_method_pictures$DualSimplex[[k]], col=grey(seq(0, 1, length = 64)), axes=0)
}

```



### Plot boxplots with distances to original matrices

```{r}
result_random$fit_objects$PGNMF <- NULL # we remove this method, since it is performing realy poor
```


#### RMSE distance for rows/columns
```{r fig.height = 5, fig.width = 20}
metric <- "rmse"
result_plots_rmse <-
  plot_metrics_box(
    res_methods = result_random$fit_objects,
    folder_to_save_results = dir_to_save_fig,
    data_used = result_random$data_used,
    title = paste0(
      metric,
      "_",
      noise_nature,
      "_random_noise_",
      noise_deviation,
      '_',
      proportions,
      '_',
      current_K ,
      'K_',
      current_N,
      'N_',
      current_M
    ),
    
    W_signif_step = 0,
    H_signif_step = 0,
    V_signif_step = 0,
    W_signif_margin_top = 0.2,
    H_signif_margin_top = 0.05,
    V_signif_margin_top = 0.05,
    H_limit = NULL,
    V_limit = NULL,
    W_limit = NULL,
    log_scale_V = T,
    log_scale_W = T,
    log_scale_H = T
  )
result_plots_rmse$plot
result_plots_rmse$summarized_plot
```


#### Pearson correlation distance for rows/columns
```{r fig.height = 5, fig.width = 20}
metric <- "pearson_loss"
result_plots_pearson <-
  plot_metrics_box(
    res_methods = result_random$fit_objects,
    folder_to_save_results = dir_to_save_fig,
    data_used = result_random$data_used,
    W_metric = metric,
    H_metric = metric,
    V_metric = metric,
    title = paste0(
      metric,
      "_",
      noise_nature,
      "_random_noise_",
      noise_deviation,
      '_',
      proportions,
      '_',
      current_K ,
      'K_',
      current_N,
      'N_',
      current_M
    ),
    W_signif_step = 0.15,
    H_signif_step = 0.15,
    V_signif_step = 0.14,
    W_signif_margin_top = 0.05,
    H_signif_margin_top = 0.05,
    V_signif_margin_top = 0.05,
    W_limit = NULL,
    H_limit = NULL,
    V_limit = NULL,
    log_scale_V = T,
    log_scale_W = T,
    log_scale_H = T
  )
result_plots_pearson$plot
result_plots_pearson$summarized_plot
```


#### Make combined plot as it is in the paper
```{r fig.height = 15, fig.width = 15}
title <- paste0(
  noise_nature,
  "_random_noise_",
  noise_deviation,
  '_',
  proportions,
  '_',
  current_K ,
  'K_',
  current_N,
  'N_',
  current_M
)


result_plot <-
  ggarrange(
    result_plots_rmse$V_plot,
    result_plots_pearson$V_plot,
    result_plots_rmse$summarized_W,
    result_plots_pearson$summarized_W,
    result_plots_rmse$summarized_H,
    result_plots_pearson$summarized_H,
    nrow = 3,
    ncol = 2,
    common.legend = T,
    legend = 'none',
    align = "hv"
  )

filename <- paste0(dir_to_save_fig, "result_plot_for_figure_", title, ".svg")
ggsave(
  filename,
  result_plot,
  width = 7.5,
  height = 15,
  device = svglite
)
result_plot
```


# Nice visualization with plotly
```{r}
result_var <- result_plots_pearson
t <- result_var$W_matrix_distances
median_df_W <- t %>%
  group_by(method, run) %>%
  summarize(average = median(.data[[metric]])) %>%
  ungroup()


t <- result_var$H_matrix_distances
median_df_H <- t %>%
  group_by(method, run) %>%
  summarize(average = median(.data[[metric]])) %>%
  ungroup()


t <- result_var$D_matrix_distances
median_df_D <- t %>%
  group_by(method, run) %>%
  summarize(average = median(.data[["diff"]])) %>%
  ungroup() 
```


## Select specific model index
```{r}
dso <- result_random$fit_objects$DualSimplex[[1]]@misc$obj
data_used <- result_random$data_used[[1]]
```


## Identify X and Omega for true coordinates
```{r}
sol_no_corr <- dso$st$solution_no_corr

W_solution <- sol_no_corr$Dv_inv_W_row
W_solution[W_solution < 0] = 0
H_solution <- sol_no_corr$H_row
H_solution[H_solution < 0] <- 0

true_V <- data_used$V[rownames(W_solution), colnames(H_solution)]
true_V_noizy <- data_used$noizy_V[rownames(W_solution), colnames(H_solution)]
true_W <- data_used$true_W[rownames(W_solution), ]
true_H <- data_used$true_H[, colnames(H_solution)]
```

```{r}
true_solution_coordinates <-  dso$get_coordinates_from_external_matrices(true_W, true_H)
X_true <- true_solution_coordinates$X
Omega_true <-  t(true_solution_coordinates$Omega)


```


## Prepare final coordinate points
```{r}
library(plotly)
dims <- dso$st$n_cell_types
## Prepare set of points
### True corners
X <- X_true
Omega <- Omega_true
X <- X[, 1:dims]
Omega <- Omega[1:dims, ]
colnames(X) <- paste0("R", c(1:dims))
rownames(Omega) <- paste0("S", c(1:dims))


### Solution corners
X_solution <- dso$st$solution_proj$X
Omega_solution <- t(dso$st$solution_proj$Omega)
X_solution <- X_solution[, 1:dims]
Omega_solution <- Omega_solution[1:dims, ]
colnames(X_solution) <- paste0("R", c(1:dims))
rownames(Omega_solution) <- paste0("S", c(1:dims))

## Just all the points
### X space
toPlotX <- as.data.frame(dso$st$proj$X)[, 1:dims]
colnames(toPlotX) <- paste0("R", c(1:dims))
### Omega space
toPlotOmega <- as.data.frame(dso$st$proj$Omega)[, 1:dims]
colnames(toPlotOmega) <- paste0("S", c(1:dims))
```


## Join solution points into geometrical shape
```{r}
get_line_between_points <- function(X_p, Omega_p) {
  n <- 2
  l <- rep(list(1:nrow(X_p)), n)
  permutations <- expand.grid(l)
  indices <- c(t(as.matrix(permutations[permutations[, "Var1"] < permutations[, 'Var2'], ])))
  X_points <- X_p[indices, ]
  Omega_points <- Omega_p[, indices]
  return(list(X_points = X_points, Omega_points = Omega_points))
}

## Prepare lines to draw between corners true
line_result <- get_line_between_points(X, Omega)
X_points <- line_result$X_points
Omega_points <- line_result$Omega_points

## Prepare lines to draw between corners true
line_result <- get_line_between_points(X_solution, Omega_solution)
X_points_solution <- line_result$X_points
Omega_points_solution <- line_result$Omega_points
```


## Plot X
```{r}
t <- list(family = "Helvetica",
          size = 30)

fig_X <-
  plot_ly(
    toPlotX,
    width = 500,
    height = 500,
    type = 'scatter3d',
    mode = 'markers'
  ) %>%
  config(toImageButtonOptions = list(
    format = "svg",
    width = 600,
    height = 600
  )) %>%
  add_trace(
    x = ~ R2,
    y = ~ R3,
    z = ~ R4,
    size = 1,
    marker = list(
      symbol = 'circle',
      color = "gray",
      opacity = 0.6,
      line = list(
        color = 'black',
        opacity = 0.4,
        width = 1
      )
    )
  ) %>%
  add_trace(
    name = "True points",
    x = X_points[, 'R2'],
    y = X_points[, 'R3'],
    z = X_points[, 'R4'],
    marker = list(color = 'orange'),
    line = list(color = 'rgb(44, 160, 44)', width = 5),
    mode = 'markers+lines'
  ) %>%
  add_trace(
    name = "Predicted points",
    x = X_points_solution[, 'R2'],
    y = X_points_solution[, 'R3'],
    z = X_points_solution[, 'R4'],
    marker = list(color = 'orange'),
    line = list(color = 'rgb(160, 0, 44)', width = 5),
    mode = 'markers+lines'
  )


fig_X <- fig_X %>% layout(showlegend = FALSE,
                          scene = list(
                            xaxis = list(title = 'R2',  titlefont = t),
                            yaxis = list(title = 'R3', titlefont = t),
                            zaxis = list(title = 'R4', titlefont = t)
                          )) %>% config(toImageButtonOptions = list(
                            format = "svg",
                            width = 600,
                            height = 600
                          ))

fig_X
```

## Plot Omega
```{r}
n <- 2

fig_Omega <-
  plot_ly(
    toPlotOmega,
    width = 500,
    height = 500,
    type = 'scatter3d',
    mode = 'markers'
  ) %>%
  config(toImageButtonOptions = list(
    format = "svg",
    width = 600,
    height = 600
  )) %>%
  add_trace(
    x = ~ S2,
    y = ~ S3,
    z = ~ S4,
    size = 3,
    marker = list(
      symbol = 'circle',
      size = 5,
      opacity = 0.6,
      color = "gray",
      line = list(color = 'black', width = 3)
    )
  ) %>%
  add_trace(
    x = Omega_points['S2', ],
    y = Omega_points['S3', ],
    z = Omega_points['S4', ],
    marker = list(color = 'orange'),
    line = list(color = 'rgb(44, 160, 44)', width = 5),
    mode = 'markers+lines'
  ) %>%
  add_trace(
    x = Omega_points_solution['S2', ],
    y = Omega_points_solution['S3', ],
    z = Omega_points_solution['S4', ],
    marker = list(color = 'orange'),
    line = list(color = 'rgb(0, 44, 160)', width = 5),
    mode = 'markers+lines'
  )


fig_Omega <- fig_Omega %>% layout(showlegend = F,
                                  scene = list(
                                    xaxis = list(title = 'S2',  titlefont = t),
                                    yaxis = list(title = 'S3', titlefont = t),
                                    zaxis = list(title = 'S4', titlefont = t)
                                  )) %>% config(toImageButtonOptions = list(
                                    format = "svg",
                                    width = 600,
                                    height = 600
                                  ))

fig_Omega
```

