---
title: "Figures 6C: Preprocessing of TCGA-HNSC for DualSimplex"
output:
  html_document:
    df_print: paged
    self_contained: yes
---

```{r}
source("../R/figure_utils.R")
source("../R/setup.R")
```



Some parameters for later
```{r}
pt_size <- 0.2
n_ct_preproc <- 20
save_dir <- "../out/dualsimplex_save_tcga_hnscc_4kg_9ct_no_hk_neighbours/"
```


# Load, preprocess and annotate data

## Data
```{r}
data_raw <- as.matrix(readRDS("../data/large/HNSC.Rda"))
# data_raw <- remove_zero_rows(data_raw)
# data_raw <- linearize_dataset(data_raw)
# data_raw <- replace_duplicate_genes(data_raw)
dim(data_raw)
```

## Anotation
```{r}
patient_data <- read.csv(
  "../data/HNSC/clinical/nationwidechildrens.org_clinical_patient_hnsc.txt",
  sep = "\t"
)
patient_data <- patient_data[3:nrow(patient_data),]
rownames(patient_data) <- patient_data$bcr_patient_barcode

sample_data <- read.csv(
  "../data/HNSC/biomedical/nationwidechildrens.org_biospecimen_sample_hnsc.txt",
  sep = "\t"
)
rownames(sample_data) <- sample_data$bcr_sample_barcode

sample_barcodes <- gsub("-\\w+-\\w+-\\w+$", "", colnames(data_raw))
patient_barcodes <- gsub("-\\w+$", "", sample_barcodes)

sample_data <- sample_data[sample_barcodes, ]
patient_data <- patient_data[patient_barcodes, ]

sample_data <- cbind(sample_data, patient_data)
rownames(sample_data) <- colnames(data_raw)

sample_data <- sample_data[, apply(sample_data, 2, function(x) { length(unique(x)) != 1 })]
sample_data <- sample_data[, grep("(_uuid|_barcode|\\.\\d)$", colnames(sample_data), invert = T)]

sample_types <- sample_data$sample_type
```

Fetch ribosome and mitochondrion pathways from msigdb for gene annotation
```{r}
# H: 50
# C2CGP: 3358
# GOCC: 1001, GO: 14765
msigdbr_df <- rbind(
  msigdbr::msigdbr(species = "Homo sapiens", category = "C5", subcategory = "GO:CC"),
  msigdbr::msigdbr(species = "Homo sapiens", category = "C5", subcategory = "GO:BP")
)
pathways <- split(x = msigdbr_df$gene_symbol, f = msigdbr_df$gs_name)
ribosome_pathways <- names(pathways)[grep("GO_.*RIBOSOM.*", names(pathways))]
mitochondrion_pathways <- names(pathways)[grep("GO_.*MITOCHOND.*", names(pathways))]
```

```{r}
gene_anno <- list(
  HK = read_gene_list("../data/gene_annotation_lists/HK.txt"),
  CC = read_gene_list("../data/gene_annotation_lists/CC.txt"),
  CODING = read_gene_list("../data/gene_annotation_lists/CODING_ENSEMBL.txt"),
  MITO = unique(unlist(pathways[mitochondrion_pathways])),
  RIBO = unique(unlist(pathways[ribosome_pathways]))
)
sample_anno <- list(ST_TUMOR = colnames(data_raw)[sample_types == "Primary Tumor"])
```

## Collapse similar gene names
```{r}
data_raw <- data_raw[unique(rownames(data_raw)),]
data_raw <- remove_zero_rows(data_raw)
data_raw <- remove_zero_cols(data_raw)

```

# DualSimplex-based preprocesing
```{r}
# Create a new Linseed 2 object. Downstream analysis will happen inside it.
lo2 <- DualSimplexSolver$new()
lo2$set_save_dir(save_dir)
```

```{r}
# To calculate this step faster, reduce the initial dataset size to 10-20k genes
# Takes: ~1 min for 55k genes
lo2$set_data(data_raw, gene_anno, sample_anno, max_sinkhorn_iterations  = 50)

# Saving memory
rm(data_raw)
```


See below, which annotations are available for genes and samples.
There are both user-defined and automatic annotations.
```{r}
# lo2$get_data() is a conventional ExpressionSet.

# Gene annotations
head(fData(lo2$get_data()))

# Sample annotations
head(pData(lo2$get_data()))
```




## MAD filtering
First we need to determine the number of cell types for deconvolution.
It is usually identified by an elbow point on the SVD plot.
However, the plot below does not have any elbow point.
That is because most of the 56k genes are not specific to any cell type,
but introduce some small irrelevant variations.
We need to filter such genes out to concentrate on what's important.
```{r}
lo2$plot_svd()
```

The first filtering step is to remove lowly varied genes.
There are quite a lot of them (see the MAD plot below), and just because
of that they take over the SVD components, although
they are meaningless in terms of cell types.
```{r}
lo2$plot_mad()
```


We can filter them out using the same MAD metric.
We care only about the right part of the plot, highly variated genes,
so theoretically, any MAD threshold lower than the right mode on the plot,
is suitable for MAD filtering. However, we found it practical
to set a low threshold at this stage, and see when the SVD plot gets better.
For this dataset, log_mad > 0.1 yields an SVD plot with a decent elbow (see below).
```{r}
# Takes: ~35s for 25k genes
lo2$basic_filter(
  log_mad_gt = 0.1,  # Note that this is log_mad. It is equivalent to mad < ~1.1
  remove_true_cols_default = FALSE,  # This function filters some gene categories by default, but for this demo we turn it off
  genes = T
)
lo2$plot_svd()
lo2$plot_mad()
```

```{r}
# Checkpoint
# We specified save_dir in the very beginning
# lo2$save_state()
# lo2 <- DualSimplexSolver$from_state(save_dir, save_there = T)
```


## Add HK neighbours annotation
We found that housekeeping genes spoil the deconvolution (see the paper),
so here is an additional step to remove them and similar genes,
using simplex projection. Housekeeping genes are naturally in the middle
of the simplex, and cell type specific genes are in the small corners.
We need only the second ones for deconvolution.

Note: Here we need to make a first projection, to operate inside a simplex
plane and deal with relations between genes.
We make the first projection, with a number of cell types, greater than
indicated by SVD, to not miss out on any relevant variance.
Judging by the plot above, 20 components is enough to capture everything before the elbow.
```{r fig.width = 14, fig.height = 6}
# Takes: ~1m for 25k genes
lo2$project(n_ct_preproc)
set.seed(3)
lo2$run_umap(neighbors_X = 10, neighbors_Omega = 10)
lo2$plot_projected(color_genes = gene_anno$HK, use_dims = NULL, pt_size = pt_size)  # NULL means default, either 2:3 or UMAP if present
```


Here again we need to find out, how many genes to filter out.
The cell type specific genes are far from the housekeeping cluster,
they are in the right part of the plot below, so we can select any threshold,
which does not remove this right part.
```{r}
# You may play around with this threshold
hk_knnd_threshold <- 0.015
log_hk_knnd_threshold <- log10(hk_knnd_threshold)
hk_genes <- gene_anno$HK[gene_anno$HK %in% rownames(lo2$st$proj$X)]
knns <- FNN::get.knnx(lo2$st$proj$X[hk_genes, ], lo2$st$proj$X, k = 10)
fData(lo2$st$data)$hk_10_dist <- knns$nn.dist[, ncol(knns$nn.dist)]
plot(sort(log10(fData(lo2$st$data)$hk_10_dist)))
abline(h = log_hk_knnd_threshold, col = "red")
```

Here are the genes, which will be filtered out using the selected threshold.
```{r fig.width = 14, fig.height = 6}
filter_out <- rownames(lo2$get_data())[fData(lo2$st$data)$hk_10_dist < hk_knnd_threshold]

# Here we set the HK_neighbours property directly on lo2, to filter them out using basic_filter,
#  which records a filtering log entry
fData(lo2$st$data)$HK_neighbours <- rownames(lo2$get_data()) %in% filter_out

lo2$plot_projected(color_genes = filter_out, use_dims = NULL, pt_size = pt_size)
```


```{r}
# Checkpoint
# We specified save_dir in the very beginning
# lo2$save_state()
# lo2 <- DualSimplexSolver$from_state(save_dir, save_there = T)
```


## Apply gene annotation filters
Now that we added HK_neighbours annotation, we can filter them out along with some other unwanted genes.

Here are the gene categories to be removed, along with the locations of the corresponding genes on the UMAP.
```{r fig.width = 14, fig.height = 6}
remove_categories <- c("RPLS", "LOC", "ORF", "SNOR", "RRNA", "RIBO", "MITO", "HK_neighbours")
for (cat in remove_categories) {
  show(lo2$plot_projected(color_genes = cat, use_dims = NULL, pt_size = pt_size))
}
```

Additionaly, only coding genes are kept.
```{r fig.width = 14, fig.height = 6}
lo2$plot_projected(color_genes = "CODING", use_dims = NULL, pt_size = 0.2)
```

```{r}
lo2$basic_filter(
  log_mad_gt = 0.1,
  remove_true_cols_default = F,
  remove_true_cols_additional = remove_categories,
  keep_true_cols = "CODING",
  genes = T
)
lo2$plot_svd()
lo2$plot_mad()
```


Here we can see that only the angles (needed for deconvolution) remain,
while the massive housekeeping middle is removed.
```{r fig.width = 14, fig.height = 6}
lo2$project(n_ct_preproc)
set.seed(4)
lo2$run_umap(neighbors_X = 10, neighbors_Omega = 10)
lo2$plot_projected(use_dims = NULL)
```


Here is what has been done so far (the package records filtering functions calls):
```{r}
lo2$get_filtering_stats()
```

```{r}
# Checkpoint
# We specified save_dir in the very beginning
lo2$save_state()
lo2 <- DualSimplexSolver$from_state(save_dir, save_there = T)
```


## Outlier filtering and finding a good projection
As you have probably noticed in the UMAP above, there are some genes
and samples, which are colored red in terms of "zero distance".
Roughly speaking that means that they are far from the middle of the simplex
by the simplex plane coordinates, and most probably, they are outliers.

Indeed, below we can see that dim_2 is occupied by some 3 genes and a single sample.
That's not what we want for deconvolution, because that is oulier variance,
not cell type variance, taking up our precious SVD components,
and shifting the simplex plane away from the actual simplex.
```{r fig.width = 14, fig.height = 6}
lo2$plot_projected(use_dims = 2:3)
```


To guide the outlier filtering process, we only need to see some first components, not all 20.
They are all contaminated by outliers.
```{r fig.width = 14, fig.height = 6}
# This will be applied to the future projection plots,
#  unless overridden explicitly with the use_dims parameter
n_ct_outliers <- 9
lo2$project(n_ct_outliers)
all_dims <- list(2:3, 4:5, 6:7, 8:9)
lo2$set_display_dims(all_dims)
lo2$plot_projected()
```


Here is the main code cell, showing everything needed to make a good projection.
High plane distance, contrary to the zero distance, means that the genes are
far from the detected simplex plane in the general N-dimensional space.
Such genes would also confuse the algorithm, which operates inside
the simplex plane, so we saw fit to filter them out too.
The plots below show:
- dataset projection onto SVD components 2-9 (R and S vectors, see the paper)
- zero_distance and plane_distance on projections
- zero_distance and plane_distance distributions (outliers are usually alone)
- zero_distance-plane_distance plot
When a sample is far from zero (from most of the samples), and close
to the plane, it means that it shifted the plane towards itself significantly.
You can see this situation in the last plot below.
```{r fig.width = 14, fig.height = 6}
lo2$project(n_ct_outliers)
lo2$plot_projection_diagnostics()
```


There are two single outlier samples, taking up dims 2-3 and others.
Let's remove them first. They are the farthest ones by zero_distance.
Note that after filtering we can't plot straight away,
we need to calculate a new projection, since we changed the data.
```{r}
lo2$distance_filter(zero_d_lt = 0.1, genes = F)
# lo2$plot_projected()  # Would raise an error
```

Just copy-paste our favorite code cell.
```{r fig.width = 14, fig.height = 6}
lo2$project(n_ct_outliers)
lo2$plot_projection_diagnostics()
```

There are still distinct outlier genes and samples.
Most probably, they are related to each other.
Let's try to filter out genes this time, and see if the outlier samples
will cease to be outliers in the absence of outlier genes.
We will remove the 3 genes, which influence dim_3 and other dimensions.
```{r fig.width = 14, fig.height = 6}
lo2$distance_filter(zero_d_lt = 0.4, genes = T)
lo2$project(n_ct_outliers)
lo2$plot_projection_diagnostics()
```


There are many red points both in samples and in genes.
In general, we will filter those points which look more alone.
Eventually, we want the farthest dim_2-dim_3 points to be red by zero_distance,
and no red in the bulk of the points.

### More iterations
Usually we do this by running one cell multiple times,
but for demonstration here we unfolded the whole process.
```{r fig.width = 14, fig.height = 6}
lo2$distance_filter(zero_d_lt = 0.2, plane_d_lt = 0.35, genes = T)
lo2$project(n_ct_outliers)
lo2$plot_projection_diagnostics()
```

```{r fig.width = 14, fig.height = 6}
lo2$distance_filter(plane_d_lt = 0.045, genes = F)
lo2$project(n_ct_outliers)
lo2$plot_projection_diagnostics()
```


```{r fig.width = 14, fig.height = 6}
lo2$distance_filter(plane_d_lt = 0.15, genes = T)
lo2$project(n_ct_outliers)
lo2$plot_projection_diagnostics()
```


```{r}
# Checkpoint
lo2$save_state()
lo2 <- DualSimplexSolver$from_state(save_dir, save_there = T)
```

## Filtering results
```{r}
lo2$get_filtering_stats()
```

```{r}
lo2$plot_svd_history(c(1, 2, 3, 8))
```

## Final projection UMAP
```{r fig.width = 11, fig.height = 5}
# Choose neighbors parameters to generate the umap with different detail level
set.seed(4)
lo2$run_umap(neighbors_X = 10, neighbors_Omega = 10)
lo2$plot_projected(use_dims = NULL, pt_size = 0.5)
```

```{r}
# Checkpoint
lo2$save_state()
# lo2 <- DualSimplexSolver$from_state(save_dir, save_there = T)
```

