---
title: "Figure 6A: Deconvolution of GSE11058"
output: html_document
---

# Imports

```{r}
# path to save everything
dir_to_save_fig <-  "../out/GSE11058/"
dir.create(file.path(".", dir_to_save_fig), showWarnings = F, recursive = T)
source("../R/figure_utils.R") # for same colors
```

## Dependencies

```{r}
library('GEOquery')
library(biomaRt)
library(hgu133plus2.db)
library(ComplexHeatmap)
library(progress)
library(plotly)
```

## Package

```{r}
unloadNamespace("linseed2")
devtools::load_all(path='../../../linseed_code/linseed2/')
```

# Prepare Data

## Get it from GEO

```{r}
n_ct <- 4

library('GEOquery')
dataset <- "GSE11058"
gse <- getGEO(dataset, AnnotGPL = T)
gse <- gse[[1]]
data_raw <- Biobase::exprs(gse)
#linearize the data!
data_raw <- linearize_dataset(data_raw)
```

## Map probes to genes

probe id will be converted and collapsed to gene names

```{r}
# probes which are present in data
probe_mapping <- biomaRt::select(hgu133plus2.db::hgu133plus2.db, rownames(data_raw), c("SYMBOL"),   keytype = "PROBEID")
probe_mapping <- probe_mapping[probe_mapping$PROBEID %in% rownames(data_raw), ]
# not empty result gene id (now we  many-to-many in in genes to probes)
probe_mapping <- probe_mapping[!is.na(probe_mapping$SYMBOL), ]
## want to collapse gene for each gene and take mean value of data
data_raw <-  data_raw[probe_mapping$PROBEID,]
rownames(data_raw) <- probe_mapping$SYMBOL
# should be unique gene name as row name after this
data_raw <-  tapply(data_raw, list(row.names(data_raw)[row(data_raw)], colnames(data_raw)[col(data_raw)]), FUN = median)

```

## Extract true proportions (H) and basis (W) from the data

Info about true samples is included for this dataset. We use this info for validation.

### True H

```{r}
pdata <- pData(gse)
component_names <-  c("Jurkat", "IM-9", "Raji", "THP-1")

mix_a <- c(2.5, 1.25, 2.5, 3.75)
mix_a <- mix_a / sum(mix_a)
mix_b <- c(0.5, 3.17, 4.75, 1.58)
mix_b <- mix_b / sum(mix_b)
mix_c <- c(0.1, 4.95, 1.65, 3.3)
mix_c <-  mix_c/ sum(mix_c)
mix_d <- c(0.02, 3.33, 3.33, 3.33)
mix_d <- mix_d / sum(mix_d)

proportions <-  matrix(0 , nrow = dim(data_raw)[[2]], ncol = n_ct )

proportions[1:3, 1] <- 1
proportions[4:6, 2] <- 1
proportions[7:9, 3] <- 1
proportions[10:12, 4] <- 1

proportions[13, ] <- mix_a
proportions[14, ] <- mix_a
proportions[15, ] <- mix_a

proportions[16, ] <- mix_b
proportions[17, ] <- mix_b
proportions[18, ] <- mix_b

proportions[19, ] <- mix_c
proportions[20, ] <- mix_c
proportions[21, ] <- mix_c

proportions[22, ] <- mix_d
proportions[23, ] <- mix_d
proportions[24, ] <- mix_d

rownames(proportions) <- rownames(pdata)
colnames(proportions) <-  component_names
true_proportions <- t(proportions)
```

### True W

```{r}
components <-  lapply(c(1:n_ct), function(comp_num){
  component_subset_columns <- colnames(true_proportions[,true_proportions[comp_num, ] == 1])
  component_subset <-  data_raw[, component_subset_columns]
  component_vector <- as.matrix(rowMedians(component_subset))
  rownames(component_vector) <- rownames(component_subset)
  return(component_vector)
  
})
true_basis <-  do.call(cbind, components)
colnames(true_basis) <-  component_names
true_marker_list <- get_signature_markers(true_basis, n_marker_genes = 100)

```

## Remove pure samples

(Optional, to make the task harder)

```{r}
#sample_is_mixed <- apply(true_proportions, 2, mad) != 0
#true_proportions <- true_proportions[, sample_is_mixed]
#data_raw <- data_raw[, sample_is_mixed]
```

## Print data dimensions

```{r}
print(paste("Dim for the whole data:", toString(dim(data_raw))))
print(paste("Dim for hidden proportions (H):", toString(dim(true_proportions))))
print(paste("Dim for hidden basis (W):", toString(dim(true_basis))))
```

## Prepare coding gene names

(optional step. these genes will be deleted from the data)

```{r}
coding_info <- biomaRt::select(hgu133plus2.db::hgu133plus2.db, rownames(data_raw),keytype="SYMBOL", c("SYMBOL", "GENETYPE"))
# possible values for genetype "protein-coding" "ncRNA""pseudo""other" "unknown" 
PROTEIN_CODING_GENES <-  coding_info[coding_info$GENETYPE == 'protein-coding', 'SYMBOL']
length(PROTEIN_CODING_GENES)
```

# Dual simplex method

## Create lo2 object and prefilter

```{r}
dso <- DualSimplexSolver$new()
dso$set_data(data_raw, gene_anno_lists=list(CODING = PROTEIN_CODING_GENES))
# remove "RPLS", "LOC", "ORF", "SNOR", keep coding only
dso$basic_filter( keep_true_cols = c("CODING"))
dso$project(n_ct)
print(paste("Dim for the prefiltered data data:", toString(dim(dso$get_data()))))
```

```{r fig.height = 5, fig.width = 11}
dso$plot_svd(dims=1:15)
dso$plot_projected("zero_distance", "zero_distance", use_dims = list(3:4))
```

## Additional filters

### MAD filtering

check distribution of median absolute deviation to see which genes will be deleted

```{r}
dummy_threshold <-  2.5
data <-  dso$get_data()
anno <-  get_anno(data)
anno$PASS_FILTER <-  FALSE
anno[anno$log_mad > dummy_threshold,]$PASS_FILTER <-  TRUE
data <- set_anno(anno, data)
plot_feature(data, feature = "log_mad", col_by ='PASS_FILTER')
```

Filter the data with selected threshold

```{r fig.height = 4, fig.width=14}
dso$basic_filter(log_mad_gt = dummy_threshold, )
dso$project(n_ct)

svd_plot_2 <- dso$plot_svd(1:10) + theme_minimal(base_size = 8)
points_2 <-  dso$plot_projected("black", "black",use_dims = 2:3, show_plots=F) +  theme_minimal(base_size = 8)
plotlist = list(points_2, svd_plot_2)
cowplot::plot_grid(plotlist=plotlist, rel_widths = c(0.66, 0.33)) 
```

Repeat previous cell with increasing thresholds to observe how SVD singular values will change. (in a perfect case should see only n_ct non-zero singular values)

```{r}
dso$plot_svd_history()
```

### Denoising the data

Plot distance of genes from the projection hyperplane

```{r fig.height = 4, fig.width=14}
distance_dist_plot_noizy <- plot_feature(dso$get_data(), feature = "plane_distance") + theme_minimal(base_size = 15)
color = fData(dso$get_data())$plane_distance
X_points_noizy <-  plot_projection_points(dso$st$proj, use_dims=2:3,color=color,color_name="plane_distance", spaces = c("X"), pt_size = 2) + theme_minimal(base_size = 15)

distances_distribution <-   plot_feature_pair(dso$get_data(), "plane_distance", "zero_distance", T, size = 0.1)


plotlist = list( X_points_noizy, distance_dist_plot_noizy, distances_distribution)

cowplot::plot_grid(plotlist=plotlist, nrow = 1, rel_widths = c(0.4, 0.3, 0.3)) 
```

Remove genes which are to far from projection hyperplane and zero point

```{r}
dso$distance_filter(plane_d_lt = 0.19, zero_d_lt = 0.3, genes = T)
dso$project(n_ct)
```

Color points by distance

```{r fig.height = 4, fig.width=14}
current_svd_plot <- dso$plot_svd(1:10) + theme_minimal(base_size = 8)
current_points_plot <-  dso$plot_projected("plane_distance", "plane_distance",use_dims = 2:3, show_plots=F) +  theme_minimal(base_size = 8)
plotlist = list(current_points_plot, current_svd_plot)
cowplot::plot_grid(plotlist=plotlist, rel_widths = c(0.66, 0.33)) 
```

Repeat previous cell with decreasing thresholds to observe how SVD singular values and coloring will change. (in a perfect case should see only n_ct non-zero singular values)

```{r}
dso$plot_svd_history()
```

## Make umap in advance for visualization purposes

since we have 3-dimensional simplex it is hard to visualize it in 2D. In practice UMAP looks good if you want to estimate the number of simplex corners and track the solution

```{r}
set.seed(23)
dso$run_umap(neighbors_X = 25, neighbors_Omega = 5)
dso$plot_projected("zero_distance", "zero_distance", use_dims = NULL)
```

## Finaly train the model

### Initialze random points

```{r}
#set.seed(72)
set.seed(33)
dso$init_solution("random")
# original space
dso$plot_projected("zero_distance", "zero_distance", use_dims = c(3:4))
# umap visualization
dso$plot_projected("zero_distance", "zero_distance", use_dims = NULL)
```

### Train multiple times with changing hyperparameters

We make learning smaller for later iterations We increase contribution of term related to positivity constraint for later iterations

```{r}

LR_DECAY_STEPS = 1
PARAMETERS_INCREASE_STEPS = 4
lr_decay <- 0.5
params_increase <- 2
original_lambda_term <- 0.01  #coef_hinge_H
original_beta_term <- 0.1 #coef_hinge_W
lr_x <-  0.001
lr_omega <- 0.001

RUNS_EACH_STEP <- 5000

pb <- progress_bar$new(total = LR_DECAY_STEPS*PARAMETERS_INCREASE_STEPS)

for  (lr_step in 1:LR_DECAY_STEPS) {
    lambda_term <-  original_lambda_term
    beta_term <- original_beta_term
      for (x in 1:PARAMETERS_INCREASE_STEPS) {
        pb$tick()
        dso$optim_solution(RUNS_EACH_STEP, 
                           optim_config(coef_hinge_H = lambda_term,
                                        coef_hinge_W =beta_term,
                                        coef_der_X = lr_x, 
                                        coef_der_Omega = lr_omega
                                        ))
        lambda_term <- lambda_term * params_increase
        beta_term <-  beta_term * params_increase

      }
    lr_x <- lr_x * lr_decay
    lr_omega <-  lr_omega * lr_decay
}
```

### Plot errors

```{r}
plot_errors(dso$st$solution_proj,   variables = c(
    "deconv_error",
    "lamdba_error",
    "beta_error",
    "total_error"
  ))
```

## Prepare plots as they are in the paper

```{r}
# This option makes all dots as a single layer reducing result image size
options("linseed-rasterize"=T)
```

### Save errors plot with rasterization

```{r}
plotErrorsWithRastr = function(metadata, variables = c("deconv_error","lamdba_error","beta_error",
"D_h_error","D_w_error","total_error")) {
  solution_proj <- metadata$st$solution_proj
  error_statistics <- solution_proj$optim_history$errors_statistics
  
  toPlot <- data.frame(error_statistics[, variables] )

  toPlot$iteration <- 0:(nrow(error_statistics)-1)
  toPlot <- melt(toPlot,id.vars="iteration",measure.vars = variables)
  plt <- ggplot(toPlot,aes(x=iteration,y=log10(value),color=variable)) +
    # rasterise(geom_point(size=0.2),dpi=600) +
     rasterise(geom_line(size=0.8), dpi=600) + theme_minimal() + labs(color="Errors")
  return(plt)
    }

errors_plot <- plotErrorsWithRastr(dso, variables = c("deconv_error",
                                 "lamdba_error",
                                 "beta_error",
                                 "total_error")) + 
  theme(text = element_text(size = 16))

filename <- paste0(dir_to_save_fig, "simulation_trajectory", ".svg")
ggsave(file=filename, plot=errors_plot, width=5, height=3, device=svg)
```

### History of solution

```{r fig.width=5, fig.height=5}
colnames(dso$st$proj$X) <- c("R1", "R2", "R3", "R4")
colnames(dso$st$proj$Omega) <- c("S1", "S2", "S3", "S4")
colnames(dso$st$proj$umap$X)  <- c("UMAP1", "UMAP2")
colnames(dso$st$proj$umap$Omega)  <- c("UMAP1", "UMAP2")

proj_solution_history <- dso$plot_projected(use_dims = (2:3), wrap=F, with_legend = F)
proj_solution_history[[1]] # genes in a space of samples
proj_solution_history[[2]] # samples in a space of genes
```

Use UMAP and decorate it how it was done in the paper

```{r}
umap_solution_history <- dso$plot_projected("zero_distance", "zero_distance", wrap=F, with_legend = F, use_dims = NULL)
pltX <- umap_solution_history[[1]] + 
  scale_fill_manual(values = colors_v[3:6]) + 
  theme_minimal(base_size=12) +
  theme(legend.position="none")
filename <- paste0(dir_to_save_fig, "proj_solution_history_X", ".svg")


ggsave(file=filename, plot=pltX, width=3, height=3, device=svg)
pltOmega <-  umap_solution_history[[2]] + 
  scale_fill_manual(values = colors_v[3:6]) +
    theme_minimal(base_size=12) +
    theme(legend.position="none")
filename <- paste0(dir_to_save_fig, "proj_solution_history_Omega", ".svg")
ggsave(file=filename, plot=pltOmega, width=3, height=3, device=svg)
ggarrange(pltX, pltOmega)
```

### Solution check plots

Get solution (W and H)

```{r}
solution <- dso$finalize_solution()
names(solution)
#solution <- lo2$get_solution()
```

#### Compare with true proportions/basis

Check ratio of true proportion or component expression to predicted

```{r fig.width=20, fig.height=5}

ptb <- coerce_pred_true_basis(solution$W, true_basis[rownames(solution$W), ])
ptp <- coerce_pred_true_props(solution$H, true_proportions)
plot_ptp_scatter(ptp)
plot_ptb_scatter(ptb)
```

#### Visualize the same as lines

```{r}
 
ptp_lines <- linseed::plotProportions(ptp[[1]], 
                                       ptp[[2]], 
                                       pnames = c("predicted", "true"),
                                       point_size = 1,
                                       line_size = 0.7
                                       ) + theme_bw(base_size=12) + theme(legend.title = element_blank(), legend.position = "bottom", axis.title.x = element_blank(), 
    axis.text.x = element_text(angle = 45, hjust = 1))

filename <- paste0(dir_to_save_fig,"ptp_lines.svg")
ggsave(filename, ptp_lines, width=3.5, height=3, device=svg)
ptp_lines
```

### Select and visualize marker genes

```{r fig.height = 5, fig.width = 11}

markers  <- dso$get_marker_genes()
to_plot <- data_raw[unlist(markers), apply(true_proportions, 2, function(x) any(x == 1))]
dso$plot_projected(rownames(to_plot), use_dims = (2:3))
```

### More decorated markers plot (as it is in the paper)

#### Visualize with UMAP

```{r}
colors <-  which_marker(rownames(fData(dso$st$data)), dso$st$marker_genes)
plot_projection_points(dso$st$proj, use_dims = NULL, spaces = c("X"), pt_size = 2, color=colors) +
  scale_color_manual(values = colors_v[3:6], na.value = adjustcolor("grey70", alpha.f = 0.7)) + 
  labs(col = "Marker Cell Type", x = "UMAP1", y = "UMAP2") +
  theme_bw(base_family = "sans", base_size = 18) +
  theme(legend.position = "right", axis.ticks = element_blank(),
        axis.text = element_blank())
```

#### Visualize original space

```{r}
colors <-  which_marker(rownames(fData(dso$st$data)), dso$st$marker_genes)
plot_projection_points(dso$st$proj, use_dims = (3:4), spaces = c("X"), pt_size = 2, color=colors) +
  scale_color_manual(values = colors_v[3:6], na.value = adjustcolor("grey70", alpha.f = 0.7)) + 
  labs(col = "Marker Cell Type", x = "UMAP1", y = "UMAP2") +
  theme_bw(base_family = "sans", base_size = 18) +
  theme(legend.position = "right", axis.ticks = element_blank(),
        axis.text = element_blank())
```

### Nice 3D visualization of markers

#### Plot genes

```{r}
library(plotly)


n <- 2
dims <-  dso$st$n_cell_types
X <-  dso$st$solution_proj$X
colnames(X) <- paste0("R", c(1:dims))
Omega <-  dso$st$solution_proj$Omega
colnames(Omega) <-  paste0("S", c(1:dims))


l <- rep(list(1:nrow(X)), n)
permutations <- expand.grid(l)
indices <-  c(t(as.matrix(permutations[permutations[,"Var1"] < permutations[, 'Var2'],] )))
X_points <- X[indices,]
Omega_points <- Omega[indices,]



colors <-  which_marker(rownames(fData(dso$st$data)), dso$st$marker_genes)

font <- list(
  family = "Helvetica",
  size = 30)
ax_settings <- list(title = '', 
                    autorange = T, 
                    showgrid = T, 
                    zeroline = F, 
                    showline = T, 
                    autotick = T, 
                    ticks = '', 
                    showticklabels = F, 
                    titlefont = font)
to_plot <- as.data.frame(dso$st$proj$X)
to_plot$color <- as.character(colors)
to_plot$color[is.na(to_plot$color)] <- "not a marker"
fig_X <- plotly::plot_ly(
  to_plot,
  x = ~R2,
  y = ~R3,
  z = ~R4,
  color = ~color,
   colors = c(colors_v[3:6], "grey70"),
  opacity = 0.9,
  size = 1,
  type = "scatter3d",
  
  mode = "markers"
)  %>%
  plotly::layout(showlegend = F,
    scene = list(
      xaxis = c(title = "R2", ax_settings),
      yaxis = c(title = "R3", ax_settings),
      zaxis = c(title = "R4", ax_settings)
    )
  )  %>%  config(toImageButtonOptions = list(format = "svg", width = 600,
                                                              height = 600))


fig_X <-  fig_X %>%  add_trace(x = X_points[,'R2'], y = X_points[,'R3'], z = X_points[,'R4'], color=NA,
            line = list(color = 'rgb(115, 115, 115)', width = 5))
fig_X
```

#### Plot samples

```{r}
font <- list(
  family = "Helvetica",
  size = 30)
ax_settings <- list(title = '', 
                    autorange = T, 
                    showgrid = T, 
                    zeroline = F, 
                    showline = T, 
                    autotick = T, 
                    ticks = '', 
                    showticklabels = F, 
                    titlefont = font)
to_plot <- as.data.frame(dso$st$proj$Omega)
fig_Omega <- plotly::plot_ly(
  to_plot,
  x = ~S2,
  y = ~S3,
  z = ~S4,
  color = "grey40",
  colors=c("navyblue"),
  opacity = 1,
  size = 1,
  type = "scatter3d",
  
  mode = "markers"
)  %>%
  plotly::layout(showlegend = FALSE,
    scene = list(
      xaxis = c(title = "S2", ax_settings),
      yaxis = c(title = "S3", ax_settings),
      zaxis = c(title = "S4", ax_settings)
    )
  )  %>%  config(toImageButtonOptions = list(format = "svg", width = 600,
                                                              height = 600))


fig_Omega <-  fig_Omega %>%  add_trace(x = Omega_points[,'S2'], y = Omega_points[,'S3'], z = Omega_points[,'S4'], color=NA,opacity = 0.4,
            line = list(color = 'rgb(115, 115, 115)',  opacity = 0.4, width = 5))
fig_Omega
```

## Plot heatmap of marker genes expressions per component

```{r fig.height = 5, fig.width = 2}
markers  <- dso$get_marker_genes()

# Gess correct order of proportions column here
solution <-  dso$finalize_solution()
reordered_H <- solution$H[guess_order(solution$H, true_proportions), ]
sorted_predicted_cell_numbers <- strtoi(gsub("^.*_", "", rownames(reordered_H)))

to_plot <- data_raw[unlist(markers), apply(true_proportions, 2, function(x) any(x == 1))]

# Minmax scaling for visualization only
to_plot <- as.data.frame(apply(to_plot, 1, function(x) (x - min(x)) / (max(x) - min(x))))



#ht_opt$TITLE_PADDING = unit(c(30, 30), "points")
#ht_opt$COLUMN_ANNO_PADDING = unit(c(20, 20), "pt")
nrows <-  dim(to_plot)[[1]]
ncols <-  dim(to_plot)[[2]]
#  "Jurkat" "IM-9"   "Raji"   "THP-1"


factor_rows <- factor(rep(rownames(true_proportions), each=nrows/n_ct), 
                      levels=c( "Jurkat", "IM-9" ,"Raji", "THP-1"))

factor_columns <- factor(rep(c(1:n_ct), each=ncols/n_ct), levels=sorted_predicted_cell_numbers)

marker_map <-  Heatmap(to_plot,name="W", 
               cluster_rows=FALSE, 
               cluster_columns = FALSE, 
               show_column_names=FALSE, 
               show_row_names=FALSE, 
               show_heatmap_legend = F, 
               column_split = factor_columns,
               row_split = factor_rows,
               use_raster =T,
               column_title_gp = gpar(fill = colors_v[3:6][sorted_predicted_cell_numbers], border = "black", fontsize =20),
               row_title_gp = gpar(fontsize =20),
               column_title_side  = "bottom"
               )
#filename <- paste0("../../figures/supplementary_benchmarks/GSE19830/heatmap.svg")
#ggsave(filename, marker_map, width=5, height=5, device=svg)
filename <- paste0(dir_to_save_fig, "heatmap.svg")
svg(filename,width=2,height=5)
draw(marker_map)
dev.off()
marker_map
```
