---
title: "R Notebook"
output: html_notebook
---
This file will produce models for S3 figure. Thisis training only, result models will be saved to out folder 

# Imports

```{r}
if (!require("NMF", quietly = TRUE))
  install.packages("NMF")
if (!require("hNMF", quietly = TRUE))
  install.packages("hNMF")
if (!require("ComplexHeatmap", quietly = TRUE))
  BiocManager::install("ComplexHeatmap")
if (!require("RcppML", quietly = TRUE))
  install.packages('RcppML')
if (!require("plotly", quietly = TRUE))
  install.packages('plotly')

```


 
```{r}
# path to save everything
dir_to_save_fig <-  "../out/NMF_simulated/"
dir.create(file.path(".", dir_to_save_fig), showWarnings = F, recursive = T)
source('../R/setup.R') # import main package
source('../R/nmf_methods.R')
```

# Simulated_data


## Params
```{r}

current_M <- 1000
current_K <- 4
current_N <- 800
```

## Single run 

### Make data and create object


```{r}

globally_generated_data <- make_simulated_data(current_M, current_N, current_K, noise_deviation =150 )

start_time <- Sys.time()
dso <- DualSimplexSolver$new()
dso$set_data(globally_generated_data$noizy_V)
dso$project(current_K)
dso$init_solution("random")
end_time <- Sys.time()
difftime(end_time, start_time, unit="mins")
```
### Check data params visualization 

```{r fig.height = 4, fig.width=14}
distance_dist_plot_noizy <- plot_feature(dso$get_data(), feature = "plane_distance") + theme_minimal(base_size = 15)
color = fData(dso$get_data())$plane_distance
X_points_noizy <-  plot_projection_points(dso$st$proj, use_dims=2:3,color=color,color_name="plane_distance", spaces = c("X"), pt_size = 2) + theme_minimal(base_size = 15)

distances_distribution <-   plot_feature_pair(dso$get_data(), "plane_distance", "zero_distance", T, size = 0.1)


plotlist = list( X_points_noizy, distance_dist_plot_noizy, distances_distribution)

cowplot::plot_grid(plotlist=plotlist, nrow = 1, rel_widths = c(0.4, 0.3, 0.3)) 
```
### Train Model


```{r}

dso$init_solution("random")
start_with_hinge_H <-  0.1
start_with_hinge_W <-  0.1
RUNS_EACH_STEP <-  2000
lr_decay <- 0.1
LR_DECAY_STEPS <-  12
PARAMETERS_INCREASE_STEPS <-  4
lr_x <-  0.1
lr_omega <-  0.1
params_increase <- 100

print(paste("current params", start_with_hinge_H, start_with_hinge_W))
original_lambda_term <- start_with_hinge_H  #coef_hinge_H
original_beta_term <- start_with_hinge_W #coef_hinge_W
original_lr_x <-  lr_x
original_lr_omega <-  lr_omega

for  (lr_step in 1:LR_DECAY_STEPS) {
  lambda_term <-  original_lambda_term * lr_x * lr_x
  beta_term <- original_beta_term   * lr_omega * lr_omega
  RUNS_EACH_STEP <-  RUNS_EACH_STEP * 0.7
  for (incr_step in 1:PARAMETERS_INCREASE_STEPS) {
    # Main training method, you can just run this
    dso$optim_solution(RUNS_EACH_STEP, 
                       optim_config(coef_hinge_H = lambda_term,
                                    coef_hinge_W =beta_term,
                                    coef_der_X = lr_x, 
                                    coef_der_Omega = lr_omega,
                                    coef_pos_D_w = 0,
                                    coef_pos_D_h = 0
                       ))
    lambda_term <- lambda_term * params_increase
    beta_term <-  beta_term * params_increase
  }
    lr_x <- lr_x * lr_decay
    lr_omega <-  lr_omega * lr_decay

  
}
solution <- dso$finalize_solution()



```



```{r}
# umap visualization
dso$plot_projected("zero_distance", "zero_distance", use_dims = NULL, with_solution = T, with_history = T, with_legend = F)
```

```{r}
dso$plot_error_history()


ggarrange(dso$plot_negative_basis_change(), dso$plot_negative_proportions_change())
```

###Try to visualize solution
(This is not mandatory code, just for your visual understanding)
Here we will plot predicted solution points together with real coordinates in a 3D space of Sinkhorn transformed matrices.

Take solution sinkhorn transformed matrices identified (H_inf and W_inf)
```{r}
sol_no_corr <-  dso$st$solution_no_corr
W_solution <- sol_no_corr$Dv_inv_W_row
W_solution[W_solution< 0] = 0
H_solution <- sol_no_corr$H_row
H_solution[H_solution < 0] <-  0
```



Take true solution matrices (Without sinkhorn transformation).

```{r}
true_V <-  globally_generated_data$V[rownames(W_solution),colnames(H_solution)]
true_V_noizy <-  globally_generated_data$noizy_V[rownames(W_solution),colnames(H_solution)]
true_W <-  globally_generated_data$true_W[rownames(W_solution),]
true_H <-  globally_generated_data$true_H[,colnames(H_solution)]
```


##### Identify coordinates in projected space for true matrices
Here we will need to perform Sinkhorn transformation for matrix V to restore D_w and D_h matrices. It can take some processing time.
We will do this for both noizy and clear matrix
<!-- true_solution_coordinates <-  dso$get_coordinates_from_external_matrices(true_W, true_H) -->
<!-- X_true <- true_solution_coordinates$X -->
<!-- Omega_true <-  t(true_solution_coordinates$Omega) -->





```{r}
R <-  dso$st$proj$meta$R
S <-  dso$st$proj$meta$S

# sinkhorn transformation for noizy matrix, which tracks all Dw and Dh matrices
extended_scaling_result <- sinkhorn_transform(true_V_noizy, data_W = true_W, data_H = true_H)

H_ss_true <-  extended_scaling_result$H_row
W_ss_true <-  extended_scaling_result$W_row
W_gs_true <-  extended_scaling_result$W_column
H_gs_true <-  extended_scaling_result$H_column

D_true <-  diag(extended_scaling_result$D_ws_column_inv[[20]])
Dh_true <-  diag(extended_scaling_result$D_hs_row_inv[[20]])

# Projection coordinates for noizy matrix components
X_true_noizy <- H_ss_true %*% t(R)
Omega_true_noizy <- S %*% W_gs_true

# sinkhorn transformation for pure matrix, which tracks all Dw and Dh matrices

extended_scaling_result <- sinkhorn_transform(true_V, data_W = true_W, data_H = true_H)

H_ss_true <-  extended_scaling_result$H_row
W_ss_true <-  extended_scaling_result$W_row
W_gs_true <-  extended_scaling_result$W_column
H_gs_true <-  extended_scaling_result$H_column

D_true <-  diag(extended_scaling_result$D_ws_column_inv[[20]])
Dh_true <-  diag(extended_scaling_result$D_hs_row_inv[[20]])

# Projection coordinates for pure matrix components
X_true <- H_ss_true %*% t(R)
Omega_true <- S %*% W_gs_true
```

##### Prepare sets of points to visualize

```{r}
library(plotly)
dims <-  dso$st$n_cell_types
## Prepare set of points 
### Noizy corners
X_noizy <- X_true_noizy
Omega_noizy <- Omega_true_noizy
X_noizy <- X_noizy[,1:dims]
Omega_noizy <- Omega_noizy[1:dims,]
colnames(X_noizy) <- paste0("R", c(1:dims))
rownames(Omega_noizy) <-  paste0("S", c(1:dims))
### True corners
X <- X_true
Omega <- Omega_true
X <- X[,1:dims]
Omega <- Omega[1:dims,]
colnames(X) <- paste0("R", c(1:dims))
rownames(Omega) <-  paste0("S", c(1:dims))


### Solution corners
X_solution <- dso$st$solution_proj$X
Omega_solution <- t(dso$st$solution_proj$Omega)
X_solution <- X_solution[,1:dims]
Omega_solution <- Omega_solution[1:dims,]
colnames(X_solution) <- paste0("R", c(1:dims))
rownames(Omega_solution) <-  paste0("S", c(1:dims))

## Just all the points
### X space
toPlotX <- as.data.frame(dso$st$proj$X)[,1:dims]
colnames(toPlotX) <- paste0("R", c(1:dims))
### Omega space
toPlotOmega <- as.data.frame(dso$st$proj$Omega)[,1:dims]
colnames(toPlotOmega) <- paste0("S", c(1:dims))

```



##### Prepare geometrical lines for plotly

```{r}

get_line_between_points <-  function(X_p, Omega_p) {
  n <- 2
  l <- rep(list(1:nrow(X_p)), n)
  permutations <- expand.grid(l)
  indices <-  c(t(as.matrix(permutations[permutations[,"Var1"] < permutations[, 'Var2'],] )))
  X_points <- X_p[indices,]
  Omega_points <- Omega_p[,indices]
  return(list(X_points=X_points, Omega_points=Omega_points))
}

## Prepare lines to draw between corners Noizy
line_result <-  get_line_between_points(X_noizy, Omega_noizy)
X_noizy_points <- line_result$X_points
Omega_noizy_points <- line_result$Omega_points

## Prepare lines to draw between corners true
line_result <-  get_line_between_points(X, Omega)
X_points <- line_result$X_points
Omega_points <- line_result$Omega_points

## Prepare lines to draw between corners true
line_result <-  get_line_between_points(X_solution, Omega_solution)
X_points_solution <- line_result$X_points
Omega_points_solution <- line_result$Omega_points

# ## Prepare lines to draw between corners corrected
# line_result <-  get_line_between_points(X_corrected, Omega_corrected)
# X_points_corrected <- line_result$X_points
# Omega_points_corrected<- line_result$Omega_points


```


```{r}
t <- list(
  family = "Helvetica",
  size = 30)

fig_X <- plot_ly(toPlotX,  width = 500, height = 500, type='scatter3d', mode='markers') %>% 
  config(toImageButtonOptions = list(format = "svg", width = 600, height = 600)) %>% 
  # add all dataset points
  add_trace( x = ~R2, y = ~R3, z = ~R4, size=1, 
             marker = list(symbol = 'circle', color = "gray", opacity = 0.8, 
                           line = list(opacity=0.4, color = 'black', width = 1))) %>%
  # add  true corner coordinates
  add_trace(x = X_points[,'R2'], y = X_points[,'R3'], z = X_points[,'R4'],  mode='markers+lines',
            line = list(color = 'rgb(44, 160, 44)', width = 5)) %>%
  # add noizy matrix corner coordinates
  add_trace(x = X_noizy_points[,'R2'], y = X_noizy_points[,'R3'], z = X_noizy_points[,'R4'],  mode='markers+lines',
            line = list(color = 'rgb(160, 0, 44)', width = 5)) %>% 
  # add solution corner coordinates
  add_trace(x = X_points_solution[,'R2'], y = X_points_solution[,'R3'], z = X_points_solution[,'R4'],  mode='markers+lines',
            line = list(color = 'rgb(0, 44, 160)', width = 5))


fig_X <- fig_X %>% layout(showlegend = FALSE, 
                          scene = list(xaxis = list(title = 'R2',  titlefont = t),

                          yaxis = list(title = 'R3', titlefont = t),

                     zaxis = list(title = 'R4', titlefont = t))
                  
                          ) %>% config(toImageButtonOptions = list(format = "svg", width = 600,
                                                              height = 600))


fig_X
```
```{r}
n <- 2
# l <- rep(list(1:nrow(X)), n)
# permutations <- expand.grid(l)
# indices <-  c(t(as.matrix(permutations[permutations[,"Var1"] < permutations[, 'Var2'],] )))
# Omega_points <- Omega[,indices]


fig_Omega <- plot_ly(toPlotOmega,  width = 500, height = 500, type='scatter3d', mode='markers') %>%  
  config(toImageButtonOptions = list(format = "svg", width = 600, height = 600)) %>%
  add_trace( x = ~S2, y = ~S3, z = ~S4, size=3, 
             marker = list(symbol = 'circle', size=5,  color = "gray", 
             line = list(color = 'black', width = 3)) ) %>% 
  add_trace(x = Omega_points['S2',], y = Omega_points['S3',], z = Omega_points['S4',],
            line = list(color = 'rgb(44, 160, 44)', width = 5),  mode='markers+lines') %>% 
  add_trace(x = Omega_noizy_points['S2',], y = Omega_noizy_points['S3',], z = Omega_noizy_points['S4',],
            line = list(color = 'rgb(160, 0, 44)', width = 5),  mode='markers+lines') %>% 
  add_trace(x = Omega_points_solution['S2',], y = Omega_points_solution['S3',], z = Omega_points_solution['S4',],
            line = list(color = 'rgb(0, 44, 160)', width = 5),  mode='markers+lines')


fig_Omega <- fig_Omega %>% layout(showlegend = FALSE,
                                  scene = list(xaxis = list(title = 'S2',  titlefont = t),
                                  yaxis = list(title = 'S3', titlefont = t),
                                  zaxis = list(title = 'S4', titlefont = t))
                  ) %>% config(toImageButtonOptions = list(format = "svg", width = 600,
                                                              height = 600))

fig_Omega
```


# Multiple runs

## Define experiment method

```{r}
run_experiment_picture_random_noizy <- function(current_K, current_M, current_N, pictures_W, 
                                                 n_data_generations = 3, 
                                                 n_runs_each_model = 10,
                                                  start_with_hinge_W=1, start_with_hinge_H=1,
                                                  LR_DECAY_STEPS = 7, RUNS_EACH_STEP = 500,    PARAMETERS_INCREASE_STEPS = 2, 
                                                  lr_x =  0.1,
                                                  lr_omega = 0.1,
                                                 noise_deviation=3.5, noise_nature="proportional" ) {
  
  data_generated <- lapply(c(1:n_data_generations), function(run_index) make_simulated_data(current_M, current_N, current_K, noise_deviation))
  total_result <- lapply(c(1:n_data_generations), function(run_index) {
      print(paste("data gen ", run_index))
      syntetic_data <- data_generated[[run_index]]
      current_res_methods <-  list()
      current_res_methods <- NMF::nmf(syntetic_data$noizy_V, current_K, list('lee', 'brunet'), .options='vp10', nrun=n_runs_each_model)

      print("Compute NMF method 'HALSacc'")
      halsacc_res <- NMF::nmf(syntetic_data$noizy_V, current_K, hNMF::HALSacc, nrun=n_runs_each_model, .options='vp10', name="HALSacc")
      #halsacc_res <- nmf(syntetic_data$V, current_K, halsacc_nmf_algorithm, nrun=10, .options='tp1', name="HALSacc")
      current_res_methods[["HALSacc"]] <- halsacc_res
      print("Compute NMF method 'PGNMF'")
      pgnmf_res <- NMF::nmf(syntetic_data$noizy_V, current_K, hNMF::PGNMF, nrun=n_runs_each_model, .options='vp10', name="PGNMF")
      # # pgnmf_res <- nmf(syntetic_data$V, current_K, pgnmf_nmf_algorithm, nrun=10, .options='tp1', name="PGNMF")
      current_res_methods[["PGNMF"]] <- pgnmf_res
      #print("Compute NMF method 'ALS'")
      als_nmf_res <- NMF::nmf(syntetic_data$noizy_V, current_K, als_nmf_algorithm, nrun=n_runs_each_model, .options='p1', name="ALS")
      current_res_methods[["ALS"]] <- als_nmf_res
      print("Compute NMF method 'DualSimplex'")
      linseed_res <- NMF::nmf(syntetic_data$noizy_V, 
                         current_K, 
                         dualsimplex_nmf_algorithm, 
                         nrun=n_runs_each_model, .options='vp10', name="DualSimplex", 
                         start_with_hinge_W=start_with_hinge_W, 
                         start_with_hinge_H=start_with_hinge_H,
                         LR_DECAY_STEPS = LR_DECAY_STEPS, 
                         RUNS_EACH_STEP = RUNS_EACH_STEP, 
                         PARAMETERS_INCREASE_STEPS=PARAMETERS_INCREASE_STEPS, 
                         lr_x=lr_x, lr_omega=lr_omega)
      current_res_methods[["DualSimplex"]] <- linseed_res
      return(current_res_methods)
  })
  
  
  all_results_flat <- unlist(total_result, recursive = F)
  unique_names <- unique(names(all_results_flat))
  per_method_per_run_result <- lapply(unique_names, function(cur_nm) all_results_flat[names(all_results_flat) == cur_nm])
  names(per_method_per_run_result) <- unique_names
 # distances_to_check <- plot_metrics(res_methods = per_method_per_run_result, data_used=data_generated, title=paste("K =", as.character(current_K), "M =", as.character(current_M),"N =",as.character(current_N)))

  
  return (list(data_used=data_generated, fit_objects=per_method_per_run_result))

}
```

## Perform experiment

```{r}
current_N  <-  300
current_M <-  200
current_K <- 4
noise_nature <-  "proportional"
# For N = 500 use   start_with_hinge_W=0.1 start_with_hinge_H=0.1
# For N = 100 use   start_with_hinge_W=10 start_with_hinge_H=1


for (noise_deviation in c(
  1
 # 0.1,
 # 0.3,  0.5, 1, 2
  )) {
  result_random <- run_experiment_picture_random_noizy(current_K = current_K, current_M= current_M, current_N = current_N, 
                                                      n_data_generations = 3, 
                                                      n_runs_each_model=3,
                                                      pictures_W = pictures_W ,  
                                                      noise_deviation=noise_deviation,
                                                      noise_nature = noise_nature,
                                                      start_with_hinge_W=1, 
                                                      start_with_hinge_H=1,
                                                      LR_DECAY_STEPS = 11,
                                                      RUNS_EACH_STEP = 2000, 
                                                      PARAMETERS_INCREASE_STEPS = 3,
                                                      lr_x =  0.1,
                                                      lr_omega = 0.1)

  print("Save to file")
  file_name <- paste0(dir_to_save_fig, noise_nature, "_random_noise_",  as.character(noise_deviation),  '_random_proportions_', 
                      current_K ,'K_', current_N, 'N_', current_M, 'M_random-H_10runs_10init_random.rds')
  saveRDS(result_random, file=file_name)


}

```




```{r}
check_triangles(result_random$fit_objects)
```


```{r, fig.height = 5, fig.width=20}
distances_to_check <- plot_metrics_box(res_methods = result_random$fit_objects, 
                                       data_used=result_random$data_used, 
                                       title = "test_simulated_result", 
                                       folder_to_save_results = dir_to_save_fig)
distances_to_check$plot 
ggarrange(distances_to_check$W_plot, distances_to_check$H_plot)

```
