---
title: "Application of DualSimplex to single cell data"
output:
  html_document:
    df_print: paged
    self_contained: yes
---

```{r}
source("../R/setup.R")
library(ggplot2)
library(Seurat)
library(systemfonts)
options("dualsimplex-rasterize" = T)
```

# Read data
```{r, fig.width=7, fig.height=5}
dso <- DualSimplexSolver$from_state("../out/single_cell_hnsc_save_3kg_16ct/")
so <- readRDS("../../single_cell/GSE103322.rds")
ct_mapping = list(
  "0" = "Cancer",
  "Fibroblast" = "Fibroblast",
  "B cell" = "B cell",
  "myocyte" = "Myocyte",
  "-Fibroblast" = "CAF",
  "Macrophage" = "Macrophage",
  "Endothelial" = "Endothelial",
  "T cell" = "T cell",
  "Dendritic" = "Dentritic",
  "Mast" = "Mast"
)
so@meta.data[, "ct"] <- unname(sapply(so@meta.data$ct, function(x) ct_mapping[[x]]))
Idents(so) <- "ct"
so_flt <- so[, colnames(so) %in% colnames(dso$get_data())]

# Prepare for clustering
so_flt <- FindVariableFeatures(so_flt, selection.method = "vst", nfeatures = 2000)
top10 <- head(VariableFeatures(so_flt), 10)

so_flt <- ScaleData(so_flt)
so_flt <- RunPCA(so_flt, features = VariableFeatures(object = so_flt))
so_flt <- FindNeighbors(so_flt, dims = 1:15)

plot1 <- VariableFeaturePlot(so_flt)
plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE)
show(plot2)

DimPlot(so_flt, reduction = "pca") + NoLegend()
ElbowPlot(so_flt)

DimPlot(so_flt)
```

```{r}
dso$plot_svd(1:100)
```


# Different projections experiments

## Project
```{r fig.width = 14, fig.height = 6}
set.seed(21)

dso$project(9)
use_dims <- NULL

dso$run_umap(neighbors_X = 20, neighbors_Omega = 15)

dso$plot_projected(
  color_genes = "zero_distance",
  color_samples = as.factor(so_flt[, colnames(dso$get_data())]$ct),
  use_dims = use_dims,
  pt_size = 0.5,
  with_legend = T,
  with_solution = F,
  with_history = F
)

plotlist <- list()
pt_size <- 0.7
plts <- dso$plot_projected(
  color_genes = "zero_distance",
  color_samples = as.factor(so_flt[, colnames(dso$get_data())]$ct),
  use_dims = use_dims,
  pt_size = pt_size,
  with_legend = T,
  with_solution = F,
  with_history = F,
  wrap = F
)
linewidth <- 0.3

plts[[2]] <- plts[[2]] + ggtitle("Cells") + theme(
  legend.position = "none",
  axis.line = element_line(color = "black", size = linewidth),
  axis.title = element_text(size = 18),
  axis.text = element_blank(),
  panel.grid = element_blank(),
  plot.title = element_text(size = 16, face = "bold"),
  text = element_text(family = "Helvetica")
) + xlab("UMAP1") + ylab("UMAP2")

plts[[1]] <- plts[[1]] + ggtitle("Genes") + theme(
  legend.position = "none",
  axis.line = element_line(color = "black", size = linewidth),
  axis.title = element_text(size = 18),
  axis.text = element_blank(),
  panel.grid = element_blank(),
  plot.title = element_text(size = 16, face = "bold"),
  text = element_text(family = "Helvetica")
) + xlab("UMAP1") + ylab("UMAP2")

plt <- cowplot::plot_grid(plotlist = plts, ncol = 2)

ggsave(
  paste0("../out/sup_sc_dims_", dso$get_n_cell_types(), ".svg"),
  plt, width = 12, height = 6, device = svglite::svglite
)

plt
```


## Solve deconvolution
### Initialize
```{r, fig.width=14, fig.height=6}
set.seed(21)
dso$init_solution("select_omega")
dso$plot_projected(use_dims = use_dims, with_legend = T)
```

### Optimize
```{r fig.width = 14, fig.height = 6}
dso$optim_solution(
  650,
  optim_config(
    coef_hinge_H = 1,
    coef_hinge_W = 1,
    coef_der_X = 0.001,
    coef_der_Omega = 0.1
  )
)
dso$plot_error_history()
dso$plot_projected(use_dims = use_dims, with_legend = T, from_iter = 1, to_iter = 650)
plot_negative_proportions_change(dso$st$proj, dso$st$solution_proj)
plot_negative_basis_change(dso$st$proj, dso$st$solution_proj)
```

```{r fig.width = 14, fig.height = 6}
lr <- 0.001
dso$optim_solution(
  2000,
  optim_config(
    coef_hinge_H = .00000001,
    coef_hinge_W = .00000001,
    coef_der_X = lr,
    coef_der_Omega = lr
  )
)
dso$plot_error_history()
dso$plot_projected(use_dims = NULL, with_legend = T)
plot_negative_proportions_change(dso$st$proj, dso$st$solution_proj)
plot_negative_basis_change(dso$st$proj, dso$st$solution_proj)
```


```{r fig.width = 14, fig.height = 6}
lr <- 0.001
dso$optim_solution(
  3000,
  optim_config(
    coef_hinge_H = .001,
    coef_hinge_W = .001,
    coef_der_X = lr,
    coef_der_Omega = lr
  )
)
dso$plot_error_history()
dso$plot_projected(use_dims = NULL, with_legend = T)
plot_negative_proportions_change(dso$st$proj, dso$st$solution_proj)
plot_negative_basis_change(dso$st$proj, dso$st$solution_proj)
```


```{r}
solution <- dso$finalize_solution()
```



## Analyze solution
### Markers and optimization
```{r fig.width=11, fig.height=5}
corner_genes <- get_signature_markers(dso$get_solution()$W, 40)
corner_cells <- lapply(
  dso$get_ct_names(),
  function(ct) {
    rownames(dplyr::slice_max(as.data.frame(t(dso$get_solution()$H)), order_by = get(ct), n = 70))
  }
)
names(corner_cells) <- dso$get_ct_names()

plts <- dso$plot_projected(
  color_genes = which_marker(rownames(dso$get_data()), corner_genes),
  color_samples = which_marker(colnames(dso$get_data()), corner_cells),
  use_dims = use_dims,
  pt_size = pt_size,
  with_legend = F,
  with_solution = T,
  with_history = T,
  color_history = F,
  wrap = F
)
linewidth <- 0.3

plts[[2]] <- plts[[2]] + ggtitle("Cells") + theme(
  axis.line = element_line(color = "black", size = linewidth),
  axis.title = element_text(size = 14),
  axis.text = element_blank(),
  panel.grid = element_blank(),
  plot.title = element_text(size = 14, face = "bold", vjust = -5, hjust = 0.02)
) + xlab("UMAP1") + ylab("UMAP2")

plts[[1]] <- plts[[1]] + ggtitle("Genes") + theme(
  axis.line = element_line(color = "black", size = linewidth),
  axis.title = element_text(size = 14),
  axis.text = element_blank(),
  panel.grid = element_blank(),
  plot.title = element_text(size = 14, face = "bold", vjust = -5, hjust = 0.02)
) + xlab("UMAP1") + ylab("UMAP2")

plt <- cowplot::plot_grid(plotlist = plts, ncol = 2)
ggsave(
  paste0("../out/sup_sc_corners_", dso$get_n_cell_types(), ".svg"), plt,
  width = 11, height = 5, device = svglite::svglite
)

plt
```


### Proportions on single cell
```{r, fig.width=20, fig.height=20}
plotlist <- list()
for (ctn in 1:dso$get_n_cell_types()) {
  ct <- dso$get_ct_names()[[ctn]]
  so_flt[[ct]] <- pmin(dso$get_solution()$H[ct, ], 0.5)
  mar <- 0
  plotlist[[ct]] <- FeaturePlot(so_flt, feature = ct, raster = T) +
    NoLegend() +
    NoAxes() +
    ggtitle(ctn) +
    theme(
      plot.title = element_text(size = 24, face = "bold", vjust = -7, hjust = 0.05),
      plot.margin = margin(t = mar, r = mar, b = mar, l = mar)
    )
}
plt <- cowplot::plot_grid(plotlist = plotlist, nrow = 4, scale = 1.05)
ggsave(
  paste0("../out/sup_sc_proportions_umap_", dso$get_n_cell_types(), ".svg"),
    plt, width = 20, height = 20, device = svglite::svglite
)
plt
```

### Proportions histograms
```{r, fig.width=12, fig.height=10}
to_plot <- reshape2::melt(t(dso$get_solution()$H))
to_plot$Var2 <- as.character(to_plot$Var2)
to_plot$Var2 <- substr(to_plot$Var2, 11, nchar(to_plot$Var2))
to_plot$Var2 <- factor(to_plot$Var2, levels = as.character(1:dso$get_n_cell_types()))
to_plot <- to_plot[to_plot$value > 0.001, ]
plt <- ggplot(to_plot, aes(x = value, fill = Var2)) +
  geom_histogram(bins = 50) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 0, vjust = 0.8, hjust=0.5, size = 10),
    # axis.title.x = element_blank(),
    axis.title.y = element_text(size = 16),
    axis.title.x = element_text(size = 16),
    legend.position= "none",
    text = element_text(family = "Helvetica", size = 12),
    panel.spacing.x = unit(2, "lines"),
    panel.spacing.y = unit(1, "lines"),
    #strip.text = elem(),
    strip.background = element_blank()
  ) +
  scale_x_continuous(breaks = c(0, 1)) +
  xlab("Cell Type Proportion") +
  ylab("Number of cells") +
  # ylim(c(0, 1000)) +
  facet_wrap(facets = vars(Var2), nrow = 4, scales = "free_y")

ggsave(
  paste0("../out/sup_sc_proportions_hist_", dso$get_n_cell_types(), ".svg"),
  plt, width = 12, height = 10, device = svglite::svglite
)

plt
```


### Proportions clustering
```{r}
dso <- DualSimplexSolver$from_state("../out/single_cell_hnsc_save_3kg_3ct/")
```

```{r fig.width=5, fig.height=5}
fn <- function(x) {
  mx <- which.max(x)
  if (x[mx] > 0) {
    return(mx)
  } else {
    return(NA)
  }
}
so_flt$dualsimplex_cluster <- apply(dso$get_solution()$H, 2, fn)[colnames(so_flt)]
plt <- (DimPlot(so_flt, group.by = "dualsimplex_cluster", label = F, raster = T) +
  NoLegend() +
  NoAxes() +
  ggtitle("")) %>% LabelClusters(id = "dualsimplex_cluster", fontface = "bold", color = "black", repel = 0.1, size = 3, box = T)

ggsave(
  paste0("../out/sup_sc_proportions_clustering_", dso$get_n_cell_types(), ".svg"),
  plt, width = 5, height = 5, device = svglite::svglite
)
plt
```

```{r}
dso$set_save_dir(paste0("../out/single_cell_hnsc_save_3kg_", dso$get_n_cell_types(), "ct"))
dso$save_state()
```


## Seurat clustering for comparison
```{r, fig.width=7, fig.height=5}
so_flt <- FindClusters(so_flt, resolution = 0.8015)
DimPlot(so_flt, label=T)
```


## Additional solution analysis
Simple negativity check
```{r}
solution_scaled <- reverse_solution_projection(dso$st$solution_proj, dso$st$proj)
sum(solution_scaled$W_col < 0) / length(solution_scaled$W_col)
sum(solution_scaled$H_row < 0) / length(solution_scaled$H_row)
plot_negative_proportions_change(dso$st$proj, dso$st$solution_proj)
plot_negative_basis_change(dso$st$proj, dso$st$solution_proj)
plot_matrix_hist_with_negative(solution_scaled$H_row, title = "Proportions distribution", bins = 100)
plot_matrix_hist_with_negative(solution_scaled$W_col, title = "Basis distribution", bins = 100)
boxplot(rowSums(solution_scaled$H_row))
boxplot(colSums(solution_scaled$W_col))
```

```{r, fig.width=20, fig.height=20}
plotlist <- list()
for (ctn in 1:dso$get_n_cell_types()) {
  ct <- dso$get_ct_names()[[ctn]]
  so_flt[[ct]] <- solution_scaled$H_row[ct, ]
  mar <- 0
  plotlist[[ct]] <- FeaturePlot(so_flt, feature = ct, raster = T) +
    NoLegend() +
    NoAxes() +
    ggtitle(ctn) +
    theme(
      plot.title = element_text(size = 24, face = "bold", vjust = -7, hjust = 0.05),
      plot.margin = margin(t = mar, r = mar, b = mar, l = mar)
    )
}
plt <- cowplot::plot_grid(plotlist = plotlist, nrow = 4, scale = 1.05)
plt
```


```{r}
# Fix negative proportions and basis
solution_scaled$H_row[solution_scaled$H_row < 0] <- 0
solution_scaled$H_row <- t(apply(solution_scaled$H_row, 1, function(x) x / sum(x)))
solution_scaled$W_col[solution_scaled$W_col < 0] <- 0
solution_scaled$W_col <- apply(solution_scaled$W_col, 2, function(x) x / sum(x))
plot_matrix_hist_with_negative(solution_scaled$H_row, title = "Proportions distribution", bins = 100)
plot_matrix_hist_with_negative(solution_scaled$W_col, title = "Basis distribution", bins = 100)
boxplot(rowSums(solution_scaled$H_row))
boxplot(colSums(solution_scaled$W_col))
```

```{r}
# res <- reverse_solution_sinkhorn(solution_scaled, dso$st$scaling)
res <- reverse_sinkhorn_c(solution_scaled$H_row,
                          solution_scaled$W_col,
                          dso$st$scaling$D_vs_row,
                          dso$st$scaling$D_vs_col,
                          dso$st$scaling$iterations)

all(is.na(res$H))
all(is.na(res$W))

R <- dso$st$proj$meta$R
S <- dso$st$proj$meta$S
V_ss <- dso$st$scaling$V_row

m <- t(S) %*% S %*% V_ss %*% t(R) %*% R
sum(m < 0) / (nrow(m) * ncol(m))
```

```{r}
plot_matrix_hist_with_negative(res$H, title = "Proportions distribution", bins = 100)
plot_matrix_hist_with_negative(res$W, title = "Basis distribution", bins = 100)

boxplot(c(res$H))
boxplot(c(res$W))
boxplot(colSums(res$H))
```


# Select Omega init, KNN KMeans clusters
```{r fig.width=7, fig.height=5}
for (nct in 11:14) {
  dso <- DualSimplexSolver$from_state(paste0("../out/single_cell_hnsc_save_3kg_", nct, "ct"))
  set.seed(21)
  dso$init_solution("select_omega")
  knns_cells <- FNN::get.knnx(dso$st$proj$Omega, dso$st$solution_proj$Omega, k = 70)
  corner_knn_cells <- as.list(as.data.frame(
    apply(knns_cells$nn.index, 1, function(x) colnames(dso$get_data())[x])
  ))
  plts <- dso$plot_projected(
    color_genes = which_marker(rownames(dso$get_data()), corner_knn_genes),
    color_samples = which_marker(colnames(dso$get_data()), corner_knn_cells),
    use_dims = use_dims,
    pt_size = pt_size,
    with_legend = F,
    with_solution = T,
    with_history = T,
    color_history = F,
    wrap = F
  )
  linewidth <- 0.3
  plts[[2]] <- plts[[2]] + ggtitle("Cells") + theme(
    axis.line = element_line(color = "black", size = linewidth),
    axis.title = element_text(size = 14),
    axis.text = element_blank(),
    panel.grid = element_blank(),
    plot.title = element_text(size = 14, face = "bold", vjust = -5, hjust = 0.02)
  ) + xlab("UMAP1") + ylab("UMAP2")
  show(plts[[2]])

  kmeans_cells_clustering <- stats::kmeans(dso$st$proj$Omega, centers = dso$st$solution_proj$Omega)
  gg_color_hue <- function(n) {
    hues = seq(15, 375, length = n + 1)
    hcl(h = hues, l = 65, c = 100)[1:n]
  }

  so_flt$marker_cell <- which_marker(colnames(so), corner_knn_cells)
  so_flt$kmeans_cluster <- kmeans_cells_clustering$cluster[colnames(so_flt)]
  show(DimPlot(
    so_flt,
    group.by = "marker_cell",
    na.value = "grey92",
    cols = gg_color_hue(dso$get_n_cell_types())
  ) + ggtitle("") + NoAxes())
  show(DimPlot(so_flt, group.by = "kmeans_cluster", label=T) + ggtitle("") + NoAxes())
}
```


```{r fig.width=7, fig.height=5}
nct <- 9
dso <- DualSimplexSolver$from_state(paste0("../out/single_cell_hnsc_save_3kg_9ct"))
set.seed(21)
dso$init_solution("select_omega")
knns_cells <- FNN::get.knnx(dso$st$proj$Omega, dso$st$solution_proj$Omega, k = 70)
corner_knn_cells <- as.list(as.data.frame(
  apply(knns_cells$nn.index, 1, function(x) colnames(dso$get_data())[x])
))
plts <- dso$plot_projected(
  color_samples = which_marker(colnames(dso$get_data()), corner_knn_cells),
  use_dims = use_dims,
  pt_size = pt_size,
  with_legend = F,
  with_solution = T,
  with_history = T,
  color_history = F,
  wrap = F
)
linewidth <- 0.3
plts[[2]] <- plts[[2]] + ggtitle("Cells") + theme(
  axis.line = element_line(color = "black", size = linewidth),
  axis.title = element_text(size = 14),
  axis.text = element_blank(),
  panel.grid = element_blank(),
  plot.title = element_text(size = 14, face = "bold", vjust = -5, hjust = 0.02)
) + xlab("UMAP1") + ylab("UMAP2")
show(plts[[2]])

kmeans_cells_clustering <- stats::kmeans(dso$st$proj$Omega, centers = dso$st$solution_proj$Omega)
gg_color_hue <- function(n) {
  hues = seq(15, 375, length = n + 1)
  hcl(h = hues, l = 65, c = 100)[1:n]
}

so_flt$marker_cell <- which_marker(colnames(so), corner_knn_cells)
so_flt$kmeans_cluster <- kmeans_cells_clustering$cluster[colnames(so_flt)]
show(DimPlot(
  so_flt,
  group.by = "marker_cell",
  na.value = "grey92",
  cols = gg_color_hue(dso$get_n_cell_types())
) + ggtitle("") + NoAxes())
show(DimPlot(so_flt, group.by = "kmeans_cluster", label=T) + ggtitle("") + NoAxes())
```


# Matching cell types
```{r}
dso_sc <- DualSimplexSolver$from_state("../out/single_cell_hnsc_save_3kg_16ct/")
dso_bulk <- DualSimplexSolver$from_state("../out/dualsimplex_save_tcga_hnscc_4kg_9ct_no_hk_neighbours/")
```

```{r fig.width = 14, fig.height = 6}
dso_sc$plot_projected()
```

```{r fig.width = 14, fig.height = 6}
dso_bulk$plot_projected()
```


```{r}
ka <- dso_bulk$get_n_cell_types()
kb <- dso_sc$get_n_cell_types()

mat <- matrix(rep(0, ka*kb), nrow = ka, ncol = kb)

for (ct1 in 1:ka) {
  for (ct2 in 1:kb) {
    a <- dso_bulk$get_marker_genes()[[dso_bulk$get_ct_names()[ct1]]]
    b <- dso_sc$get_marker_genes()[[dso_sc$get_ct_names()[ct2]]]
    iou = length(intersect(a, b)) #/ length(union(a, b))
    mat[ct1, ct2] <- iou
  }
}
rownames(mat) <- dso_bulk$get_ct_names()
colnames(mat) <- dso_sc$get_ct_names()
pheatmap::pheatmap(mat, display_numbers = T)
```

```{r}
cancer_bulk <- c(4, 5, 6, 8, 9)
cancer_sc <- c(5, 11, 12, 2, 8)

pheatmap::pheatmap(mat[cancer_bulk, cancer_sc], display_numbers = T)
```


```{r}
ka <- dso_bulk$get_n_cell_types()
kb <- dso_bulk$get_n_cell_types()

mat <- matrix(rep(0, ka*kb), nrow = ka, ncol = kb)

for (ct1 in 1:ka) {
  for (ct2 in 1:kb) {
    a <- dso_bulk$get_marker_genes()[[dso_bulk$get_ct_names()[ct1]]]
    b <- dso_bulk$get_marker_genes()[[dso_bulk$get_ct_names()[ct2]]]
    iou = length(intersect(a, b)) #/ length(union(a, b))
    mat[ct1, ct2] <- iou
  }
}
rownames(mat) <- dso_bulk$get_ct_names()
colnames(mat) <- dso_bulk$get_ct_names()
pheatmap::pheatmap(mat, display_numbers = T)
```

```{r}
ka <- dso_sc$get_n_cell_types()
kb <- dso_sc$get_n_cell_types()

mat <- matrix(rep(0, ka*kb), nrow = ka, ncol = kb)

for (ct1 in 1:ka) {
  for (ct2 in 1:kb) {
    a <- dso_sc$get_marker_genes()[[dso_sc$get_ct_names()[ct1]]]
    b <- dso_sc$get_marker_genes()[[dso_sc$get_ct_names()[ct2]]]
    iou = length(intersect(a, b)) #/ length(union(a, b))
    mat[ct1, ct2] <- iou
  }
}
rownames(mat) <- dso_sc$get_ct_names()
colnames(mat) <- dso_sc$get_ct_names()
pheatmap::pheatmap(mat, display_numbers = T)
```


```{r fig.width=20, fig.height=15}
so_flt <- add_list_markers(so_flt, dso_bulk$get_marker_genes())
plot_marker_enrichment(so_flt, dso_bulk$get_ct_names(), limits = NULL)
```

```{r fig.width=20, fig.height=20}
so_flt <- add_list_markers(so_flt, dso_sc$get_marker_genes())
plot_marker_enrichment(so_flt, dso_sc$get_ct_names(), limits = NULL)
```

