---
title: "R Notebook"
output: html_notebook
---

# Repeat of partial figure GSE11058
Here we assume that data is already generated (see another file)
You need to install CellMix to use dsa

## Tricky installation

Note: maximum R version supported by CellMix is 3.6. 
(docker for this https://hub.docker.com/r/aladyevae/cellmix)
Original package is not available for later version of R
But we found a fork of this repo that is working
```{r}
# Before: copy CellMix folder we provided in our repo to .libPaths() folder
if (!require("CellMix", quietly = TRUE))
  devtools::install_github('denklewer/cellmix')
library("CellMix")
library(digest)
library(matrixStats)
library(dplyr)
```


# Make Mixed Dataset



```{r}
if (!require("GEOquery", quietly = TRUE))
  BiocManager::install('GEOquery')
# BiocManager::install('Metrics')
# BiocManager::install('uwot')
#BiocManager::install('tidyverse')
```

```{r}
library(CellMix)
library(digest)
library(GEOquery)
```

```{r}
dir_to_save <-  "../data/large/signature_based/GSE11058_benchmark/"
DS_NAME <- "GSE11058"
n_markers <- 20
```

```{r}
GSE19830.em <- ExpressionMix("GSE11058", verbose=TRUE)
```

```{r}
props <- coef(GSE11058.em)

ex <- exprs(GSE11058.em)
bas <- basis(GSE11058.em)

ex <- ex[log2(apply(ex, 1, mad)) > 8, ]
bas <- bas[rownames(ex),]

cat(dim(ex), "\n")

data_full <- ex
colnames(bas) <- gsub(" |-", "_", colnames(bas))
```

## Save it to use in another R version

```{r}
saveRDS(GSE11058.em, paste0(dir_to_save, "single_run_em.rds"))
saveRDS(bas, paste0(dir_to_save, "single_run_bas.rds"))
saveRDS(props, paste0(dir_to_save, "single_run_props.rds"))
saveRDS(data_full, paste0(dir_to_save, "/single_run_data_full.rds"))

```

## Multiple runs

```{r}
for (rep_n in seq(1,5)) {
  set.seed(rep_n)
  DS_NAME <- "GSE11058"
  GSE11058.em <- ExpressionMix("GSE11058", verbose=TRUE)
  props <- coef(GSE11058.em)
  ex <- exprs(GSE11058.em)
  bas <- basis(GSE11058.em)
  ex <- ex[sample(nrow(ex), nrow(ex)*0.8), ]
  ex <- ex[log2(apply(ex, 1, mad)) > 8, ]
  bas <- bas[rownames(ex),]
  data_full <- ex
  colnames(bas) <- gsub(" |-", "_", colnames(bas))
  rownames(props) <- gsub(" |-", "_", rownames(props))
  saveRDS(GSE11058.em, paste0(dir_to_save, rep_n,"_run_em.rds"))
  saveRDS(bas, paste0(dir_to_save, rep_n,"_run_bas.rds"))
  saveRDS(props, paste0(dir_to_save, rep_n, "_run_props.rds"))
  saveRDS(data_full, paste0(dir_to_save, rep_n, "_run_data_full.rds"))
}
```



