---
title: "R Notebook"
output: html_notebook
---

# Imports


```{r}
if (!require("NMF", quietly = TRUE))
  install.packages("NMF")
if (!require("hNMF", quietly = TRUE))
  install.packages("hNMF")
if (!require("ComplexHeatmap", quietly = TRUE))
  BiocManager::install("ComplexHeatmap")
if (!require("RcppML", quietly = TRUE))
  install.packages('RcppML')
if (!require("data.table", quietly = TRUE))
  install.packages('data.table')

if (!require("ggsignif", quietly = TRUE))
  install.packages('ggsignif')
if (!require("GEOquery", quietly = TRUE))
  BiocManager::install("GEOquery")

if (!require("linseed", quietly = TRUE))
  devtools::install_github("ctlab/linseed")

if (!require("gridExtra", quietly = TRUE))
  install.packages('gridExtra')


if (!require("ggpubr", quietly = TRUE))
  install.packages('ggpubr')

if (!require("svglite", quietly = TRUE))
  install.packages('svglite')
```


 
```{r}
dir_to_save_fig <-  "../../out/NMF/" # path to save everything
dir.create(file.path(".", dir_to_save_fig), showWarnings = F, recursive = T) 
unloadNamespace("linseed2")   
devtools::load_all(path='../../../linseed2/')  # import main package

source('nmf_methods.R')  # nmf methods 
```

# Syntetic experiment 
Here we use Syntetic data from NMF package just to check that everything is working fine
(to visualize and check that everything work)
## Make the NMF Data matrix

```{r}
components_count <-  c(3, 4, 5)
mixtures_count <-  c(25, 50, 100, 500)
features_count <- c(200, 500, 1000, 2000)

```

```{r}
current_K <- components_count[[2]]
current_M <- features_count[[2]]
current_N <- mixtures_count[[1]]
```


```{r}
global_generated_data <- make_syntetic_data(M=current_M, N=current_N, K=current_K)
```


```{r}
plot_w_h_v(global_generated_data$true_W, global_generated_data$true_H, global_generated_data$V)
```


## Solve this NMF using NMF package (3 different algorithms here)

```{r}
res_methods <- NMF::nmf(global_generated_data$V, current_K, list('lee', 'brunet', 'nsNMF'), nrun=5, .options='kt')
```

```{r}
res_halsAcc <-  NMF::nmf(global_generated_data$V, current_K, hNMF::HALSacc, nrun=5, .options='tk', name="HALSacc")
res_methods[["HALSacc"]] <- res_halsAcc
res_PGNMF <-  NMF::nmf(global_generated_data$V, current_K, hNMF::PGNMF, nrun=5, .options='tk', name="PGNMF")
res_methods[["PGNMF"]] <- res_PGNMF
res <-  NMF::nmf(global_generated_data$V, current_K, als_nmf_algorithm, nrun=5, .options='tk', name="ALS")
res_methods[["ALS"]] <- res
```



# Solve the same NMF problem using Dual Simplex Method

```{r}
res_methods[["Linseedv2 NMF"]] <- NULL
```

```{r}
res <-  NMF::nmf(global_generated_data$V, current_K, dualsimplex_nmf_algorithm, nrun=5, .options='tk', name="DualSimplex")
res_methods[["DualSimplex"]] <- res
```


## Check results
### Triangles found by linseed
```{r}
check_triangles(res_methods)
```

### Matrices for each algorithm
Plot W, H , V for each algorithm used
```{r}
lapply(res_methods, function (method_result_list){
  method_result <- method_result_list[[1]] 
  current_W <- method_result@fit@W
  current_H <- method_result@fit@H
  plot_w_h_v(current_W, current_H, current_W %*% current_H, method_result@method)
  
})
```
### RMSE errors


```{r, fig.height = 5, fig.width=20}
data_used <- rep(list(global_generated_data), 10)
distances_to_check <- plot_metrics_box(res_methods = res_methods, data_used=data_used, folder_to_save_results = dir_to_save_fig)
distances_to_check$plot
```

